---
title: "Lesson 3 R Activity"
author: "Rick Davila"
date: "4/26/2020"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

### Install packages

```{r setup, include=TRUE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(e1071)
library("xlsx")
library(xtable)

rm(list = ls())
```

### Read in rocket propellant data

Copy and paste the following information to  your completed Lesson 1 R Activity .R file.  Read data from Excel spreadsheet using the read.xlsx() command

```{r, , warning = FALSE}
# uncomment for laptop 1
ex2_1 <- read.xlsx("data-ex-2-1.xlsx",
                   sheetIndex = 1, 
                   colIndex = c(2,3), 
                   as.data.frame = TRUE, 
                   header = TRUE)
```

### Rocket propellant data - table printout

```{r, results="asis"}

xtable(ex2_1)

```

### Data structure and dimensions of data

Output data structure and dimensions using the str() and dim() commands

```{r}
# output dataframe structure
str(ex2_1)

# dim of data 'matrix' (i.e., should be 20 rows by 2 columns)
dim(ex2_1)
```

### Print out tabel using revised column names

```{r, results="asis"}

names(ex2_1) <- c("Shear_Strength", "Age")
attach(ex2_1)

#Printout revised table using new column names
xtable(ex2_1)

```

### Create scatter plot

```{r}

# Create scatterplot using plot()
plot(Age, Shear_Strength)

```

### Create linear model "by hand"

First, perform intermediary calculations - Syy, Sxx, Sxy, x_bar, y_bar, num

```{r}

# num = length(Age)
num = length(Age)

Syy = sum(Shear_Strength^2) - (sum(Shear_Strength))^2/num
print(sprintf("Syy = %f",Syy))

# From Eq. (2.9) in the e-book
Sxx = sum(Age^2) - (sum(Age))^2/num
print(sprintf("Sxx = %f",Sxx))

# From Eq. (2.10) in the e-book
Sxy=sum(Age*Shear_Strength)-sum(Age)*sum(Shear_Strength)/num
print(sprintf("Sxy = %f",Sxy))

#x_bar and y_bar ...
x_bar = sum(Age)/num
y_bar = sum(Shear_Strength)/num
print(sprintf("x_bar = %f and y_bar = %f",x_bar, y_bar))

```

### Determine estimates for slope and intercept

```{r}

# slope, from Eq. (2.11) in the e-book
beta_1 = Sxy/Sxx
print(sprintf("beta_1 = %f",beta_1))

# intercept, from Eq. (2.6) in the e-book
beta_0 = y_bar - beta_1*x_bar
print(sprintf("beta_0 = %f",beta_0))

print('The Least Squares regression line by hand:')

print(sprintf(" y_hat = %f + (%f)x",beta_0, beta_1))

```

The equation is 
$$ \hat{y} = (`r round(beta_1,2)`)x + (`r round(beta_0,2)`)$$

### Output the model 'by hand'

```{r, results="asis"}
# Output 'by hand' model by hand
y_hat = beta_0 + beta_1*Age
Results = data.frame(Shear_Strength, y_hat)
names(Results) <- c("Observed Values","Fitted Values")
xtable(Results)

```

### Create the linear model using lm() command and display using the summary() command

```{r, results="asis"}

# Create linear model using lm() command and display output using the summary() command
model=lm(Shear_Strength~Age)
summary(model)

```

### Example 2.2 (p.21-22)

Obtain ANOVA table elements (by hand) - SST, SS_Regression, SS_Residual

```{r}
SST = sum((Shear_Strength - y_bar)^2)
SS_Regression = sum((y_hat - y_bar)^2) 
SS_Residual = sum((Shear_Strength - y_hat)^2)

print(sprintf("note, Syy = SST = %f and SS_Reg+SS_Res = %f",Syy, SS_Regression+SS_Residual))

# Define degrees of freedom (by hand)
# num = length(Age) - number of observations
df_SST = num - 1          # 20 - 1 = 19
df_SS_Regression = 1              # 1
df_SS_Residual = num - 2  # 20 - 2 = 18

# Obtain residual degrees of freedom automatically using df.residual()
dfresidual = df.residual(model)
print(sprintf("residual degrees of freedom using df.residals = %i",dfresidual))

# Obtain estimated error variance (MS_Residual)
MS_Residual = SS_Residual/dfresidual
print(sprintf("sigma^2, estimated error variance = %f",MS_Residual))

#Obtain ANOVA table using aov() command or anova() command and display output
aov(model)
anova(model)

```

### Example 2.3 (p.25)

Test significance of slope parameter (by hand) using the qt() and pt() commands for critical value and p-value for the t-distribution

```{r}

se_beta1 = sqrt(MS_Residual/Sxx)
test_statistic = beta_1/se_beta1

# for alpha = 0.05
alpha = 0.05
tcritical_value = abs(qt(alpha/2.0, df = df_SS_Residual))
p_value = 2*pt(-abs(test_statistic), df = df_SS_Residual)
print(sprintf("test statistic = %f, critical value = %f and p-value = %e",
              test_statistic, tcritical_value, p_value))

```

The absolute value of the test statistic is greater than the critical value $$ 12.8 > 2.1 $$ and p value is $$1.64 \times 10^{-10}$$... there's a linear relationship 
between the shear strength and the age of the propellant

### Example 2.4 (p. 28)

Test for significance of Regression (F-test) at .01 significance level

(1) By hand

```{r}

# By hand:
# F_o test statistic
siglevel = 0.01

F_o = (SS_Regression/df_SS_Regression)/(SS_Residual/df_SS_Residual)
print(sprintf("F_o test statistic = %f", F_o))

Fcritical = qf(1-siglevel, df_SS_Regression, df_SS_Residual)
print(sprintf("F statistic = %f",Fcritical))

# as a check, from table A.4 in the e-book, F_(0.01,1,18) = 8.29
p_val = 1 - pf(F_o, df_SS_Regression, df_SS_Residual)
print(sprintf("F p value = %e",p_val))

```

Since $$F_0 = 165.36 > F_{(0.01,1,18)}$$ We can reject the null hypotheses, there's a relationship between the shear strength and the age of the rocket propellant


2) Using aov and anova commands

Using aov and anova functions:

aov model output

```{r}

aov(model)

```

anova model output

```{r}

anova(model)

```

### Example 3 2.5 (p 30)

Construct a $95\%$ confidence interval (CI) for the slope parameter (by hand)
The $100\left(1-\alpha\right)$ percent confidence interveral (CI) of the slope $\beta_1$ is given by
$$\hat{\beta}_1 - t_{\alpha/2,n-2}se(\hat{\beta}_1) \le \hat{\beta}_1 \le \hat{\beta}_1 + t_{\alpha/2,n-2}se(\hat{\beta}_1)$$
We construct a $95%$ CI on $\beta_1$ by using the standard error of $\hat{\beta}_1$ is $se(\hat{\beta}_1) = \sqrt{MS_{Res}/S_{xx}}$ and $t_{0.025,18} = `r tcritical_value`$

```{r}
# Example 2.5 (p. 30)
# Construct a 95% confidence interval for the slope parameter (by hand)

se_beta_1 = sqrt(MS_Residual/Sxx)
CIleft_wing = beta_1 - tcritical_value * se_beta_1
CIright_wing = beta_1 + tcritical_value * se_beta_1

```
The $95\%$ CI on the slope is
$$`r round(beta_1,2)` - (`r round(tcritical_value,2)`)(`r round(se_beta_1,2)`) \le \hat{\beta}_1 \le `r round(beta_1,2)` + (`r round(tcritical_value,2)`)(`r round(se_beta_1,2)`) $$ or


$$`r CIleft_wing` \le \hat{\beta}_1 \le`r CIright_wing`$$

check using the Confint command

```{r}

# Check using Confint command
confint(model, level = 0.95)

```

Construct 95% C.I. for the point x0 = 13.3625

Consider finding a $95\%$ CI on $E(y|x_0)$.  The CI per Eq. (2.43) in the e-book is
$$\hat{\mu}_{y|x_0} - t_{\alpha/2,n-2}\sqrt{MS_{Res}\left(\frac{1}{n} + \frac{(x_0-\bar{x})^2}{Sxx}\right)} \le E(y|x_0) \le \hat{\mu}_{y|x_0} + t_{\alpha/2,n-2}\sqrt{MS_{Res}\left(\frac{1}{n} + \frac{(x_0-\bar{x})^2}{Sxx}\right)}$$
for the Rocket Propellant data, we have
$$\hat{\mu}_{y|x_0} - (`r round(tcritical_value,2)`)\sqrt{`r round(MS_Residual,2)`\left(\frac{1}{`r num`} + \frac{(x_0-`r round(x_bar,2)`)^2}{`r round(Sxx,2)`}\right)} \le E(y|x_0) \le \hat{\mu}_{y|x_0} + (`r round(tcritical_value,2)`)\sqrt{`r round(MS_Residual,2)`\left(\frac{1}{`r num`} + \frac{(x_0-`r round(x_bar,2)`)^2}{`r round(Sxx,2)`}\right)}$$

For this case, $\hat{\mu}_{y|x_0=13.3625}$ is

```{r}

x0 = 13.3625
mu_hat_x0 = beta_1 * x0 + beta_0
mu_hat_x0

```
and the CI at $\hat{\mu}_{y|x_0=13.3625} = `r mu_hat_x0`$ is

```{r}

Sqrt_quantity = sqrt(MS_Residual*(1/num + (x0 - x_bar)^2/Sxx))
CIx0_left_wing = mu_hat_x0 - tcritical_value * Sqrt_quantity
CIx0_right_wing = mu_hat_x0 + tcritical_value * Sqrt_quantity
CIx0_left_wing
CIx0_right_wing

```
or
$$`r CIx0_left_wing` \le E(y|x_0=13.3625) \le `r CIx0_right_wing`$$

### Example 2.7 (p. 34-35)
Construct a $95\%$ prediction interval for $x_0 = 10 \text{ weeks}$

The $100(1 - \alpha)$ percent prediction interval on a future observation at $x_0$ is
$$\hat{y_0} - t_{\alpha/2,n-2}\sqrt{MS_{Res}\left(1+ \frac{1}{n} + \frac{(x_0-\bar{x})^2}{Sxx}\right)} \le y_o \le \hat{y_0} + t_{\alpha/2,n-2}\sqrt{MS_{Res}\left(1 + \frac{1}{n} + \frac{(x_0-\bar{x})^2}{Sxx}\right)}$$
for $x_0 = 10 \text{ weeks}$, $\hat{y_0}$ is
```{r}

x0 = 10
y_hat = beta_1 * x0 + beta_0
y_hat

```

$\hat{y_0} = `r y_hat`$

so, the interval becomes

$$`r round(y_hat,2)` - (`r round(tcritical_value,2)`)\sqrt{`r round(MS_Residual,2)`\left(1+ \frac{1}{`r num`} + \frac{(`r x0`-`r round(x_bar,2)`)^2}{`r round(Sxx,2)`}\right)} \le y_0 \le `r round(y_hat,2)` + (`r round(tcritical_value,2)`)\sqrt{`r round(MS_Residual,2)`\left(1 + \frac{1}{`r num`} + \frac{(`r x0`-`r round(x_bar,2)`)^2}{`r round(Sxx,2)`}\right)}$$
which simplies to
```{r}

x0 = 10
Sqrt_quantity = sqrt(MS_Residual*(1 + 1/num + (x0 - x_bar)^2/Sxx))
PI_left_wing = y_hat - tcritical_value * Sqrt_quantity
PI_right_wing = y_hat + tcritical_value * Sqrt_quantity
PI_left_wing
PI_right_wing

```

$$`r PI_left_wing` \le y_0 \le `r PI_right_wing`$$

Superimpose a $95\%$ prediction and confidence interval plots onto the scatterplot of data using predict(), plot(), lines() and order() commands.

```{r}

# Obtain prediction interval lines automatically using predict()
# sorting data frame by Age ... recall from Lesson 1, attach(ex2_1) dataset

sorted_data <- ex2_1[order(Age),]
ci_band = predict(model, sorted_data, interval = "confidence", level = 0.95)

# Obtain confidence interval lines automatically using predict()
pi_band = predict(model, sorted_data, interval = "prediction", level = 0.95)

# Create scatter plot using plot()
plot(Age, Shear_Strength)

# Add line of predicted y_hat values using lines(). 
# Be sure to use order() to ensure that x-values are in ascending (versus random) order

# sorted Age column, or x-values, in ascending order are now , contained in sorted_data[,2]
y_hat = beta_1*sorted_data[,2] + beta_0
lines(sorted_data[,2], y_hat, lty = 1)

# Add prediction interval lines using lines() and order() commands
lines(sorted_data[,2], ci_band[,2], lty = 2)
lines(sorted_data[,2], ci_band[,3], lty = 2)


# Add confidence interval lines using lines() and order() commands
lines(sorted_data[,2], pi_band[,2], lty = 3)
lines(sorted_data[,2], pi_band[,3], lty = 3)

legend("topright", legend = c("Fit","95% CI","95% PI"), lty = c(1,2,3))

```
