---
title: "Lesson 7 R Activity 2"
author: "Rick Davila"
date: "5/10/2020"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

## Lesson 7, Windmill Example - Install packages
#### Install necessary packages using library()
#### Perform data housekeeping - upload, name columns, display to make sure it reads properly, etc. 

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)

library(e1071)
library(xtable)
library("xlsx") # Needed to read data
library(psych) # For geometric mean in Example 5.3

rm(list = ls())

```

#### Read data file (data-ex-5-2.xlsx)

```{r}

ex5_2 <- read.xlsx("data-ex-5-2.xlsx",
                   sheetIndex = 1, 
                   colIndex = c(2,3), 
                   as.data.frame = TRUE, 
                   header = TRUE)

```

#### Assign labels to data columns using names() and attach() commands

```{r}

names(ex5_2) <- c("Wind_Velocity", "DC_Output")
attach(ex5_2)

```

#### Output data to make sure it reads properly

```{r results="asis"}

out <- as.data.frame(c(ex5_2))
colnames(out) <- c("Wing_Velocity", "DC_Output")
tab <- (xtable(out, digits=c(0,2,3)))
print(tab, type="html")

# Output data structure and dimensions
str(ex5_2)
dim(ex5_2)

```

## Example 5.2 (p. 178-181)
#### Create scatterplot to visualize data

```{r}

plot(Wind_Velocity,DC_Output, main = "Scatter plot of DC output vs Wind Velocity (mph)")

```

#### Fit linear model first; compare results to p. 178


```{r}

# Obtain regression estimates using lm command
model <- lm(DC_Output ~ Wind_Velocity)
beta_0 <- model$coefficients[1]
beta_1 <- model$coefficients[2]

# show beta estimates, std. error, tvalues and p-values
xtable(model)
xtable(summary(aov(model)))

```

The equation for the linear regression model: 
$$\hat{y} = (`r beta_0`) + (`r beta_1`)x$$

```{r}

# ANOVA table
# summary(aov)
out <- anova(model)

xtable(out)

```

#### Create residual plots - R-Student and raw residuals versus fitted values

```{r}

# DC_Output, given Wind_Velocity
model <- lm(DC_Output ~ Wind_Velocity)
beta_0 <- model$coefficients[1]
beta_1 <- model$coefficients[2]

# Calculate y_hat
y_hat <- beta_0 + beta_1*Wind_Velocity

# Calculate R-Student residuals
residuals <- rstudent(model)

plot(y_hat, residuals, main='Plot of R-student values t_i versus fitted values y_hat', xlab = 'y_hat', ylab = 'Rstudent() Residuals (t_i)')

```

#### Residual plots show clear nonlinearity. Attempt to rectify through transformation: x' = 1/x

```{r}

Wind_Velocity_t <- 1/Wind_Velocity

```

#### Create scatterplot and model using transformed data; compare to p. 180 and 181

```{r}

plot(Wind_Velocity_t,DC_Output, main = "Scatter plot of DC output vs Wind Velocity Transformed (mph)")

# create lm() model with transformed wind velocity
model_t <- lm(DC_Output ~ Wind_Velocity_t)

# 
xtable(model_t)
xtable(summary(aov(model_t)))

```

#### Plot residuals versus fitted values; compare to p. 182

```{r}

# Calculate r_student residuals
R_Student_Residuals_t <- rstudent(model_t)

y_hat_t <- model_t$fitted.values

plot(y_hat_t, R_Student_Residuals_t, ylab = "R-Student Residuals", xlab = "Transformed Wind Velocity", main = "R−Student Residuals versus Fitted Values")

```

## Example 5.4 (p. 187)

#### Determine alpha value to transform independent variable using Box-Tidwell Procedure
#### Step 1.0: Define alpha_0 = 1 and run simple regression model (performed in Example 5.1 above)

```{r}

# Obtain regression estimates using lm command, for alpha_0 = 0
model <- lm(DC_Output ~ Wind_Velocity)
beta_0 <- model$coefficients[1]
beta_1 <- model$coefficients[2]

# show beta estimates, std. error, tvalues and p-values
xtable(model)
xtable(summary(aov(model)))

```

For step 1.0 we have, the equation for the linear regression model is:

$$\hat{y} = \hat{\beta_0}+\hat{\beta_1}x = (`r beta_0`) + (`r beta_1`)x$$


#### Step 1.1: Define 2 = x ln x and add to model as an independent variable; compare output to p. 187


```{r}

# Then defining w = x ln x, we fit Eq. (5.8) and obtain
w <- Wind_Velocity*log(Wind_Velocity)           #  w = xlog(x)
model <- lm(DC_Output ~ Wind_Velocity + w)
beta_0_star <- model$coefficients[1]
beta_1_star <- model$coefficients[2]
gamma_1 <- model$coefficients[3]

# show beta estimates, std. error, tvalues and  p-values
xtable(model)
xtable(summary(aov(model)))

```


For step 1.1 we have, the equation for the linear regression model is:
$$\hat{y} =\hat{\beta_0}^* +\hat{\beta_1}^*x + \hat{\gamma}w= (`r beta_0_star`) + (`r beta_1_star`)x + (`r gamma_1`)w$$


#### Step 1.2: Using  Equation 5.10, calculate alpha_1; compare to p. 187 


```{r}

alpha_1 <- (gamma_1/beta_1) + 1

```


For step 1.2, we have:
$$\alpha_1 = \frac{\hat{\gamma}}{\hat{\beta_1}} + 1=(`r alpha_1`)$$

#### Step 2.0: Define x' = x^alpha_1 and regress model with alpha_0 = 0; compare to p. 187


```{r}

# Obtain regression estimates using lm command, for alpha_0 = 1
Wind_Velocity_t1 <- Wind_Velocity^(alpha_1)
model <- lm(DC_Output ~ Wind_Velocity_t1)
beta_0_t <- model$coefficients[1]
beta_1_t <- model$coefficients[2]

# show beta estimates, std. error, tvalues and  p-values
xtable(model)
xtable(summary(aov(model)))

```


For step 2.0, $x^{\prime}=x^{`r alpha_1`}$ and the equation for the linear regression model is:
$$\hat{y} = \hat{\beta_0}+\hat{\beta_1}x^{\prime} = (`r beta_0_t`) + (`r beta_1_t`)x^{\prime}$$


#### Step 2.1: Define w' = x' ln x', add as independent variable, and run model


```{r}

# Obtain regression estimates using lm command 
w_t1 <- Wind_Velocity_t1*log(Wind_Velocity_t1) # w = x'log(x')
model <- lm(DC_Output ~ Wind_Velocity_t1 + w_t1)
beta_0_star_t1 <- model$coefficients[1]
beta_1_star_t1 <- model$coefficients[2]
gamma_1_t1 <- model$coefficients[3]

# show beta estimates, std. error, tvalues and  p-values
xtable(model)
xtable(summary(aov(model)))


```


For step 2.1 we have, the equation for the linear regression model is:

$$\hat{y} =\hat{\beta_0}^* +\hat{\beta_1}^*x^{\prime}+\hat{\gamma}w^{\prime} = (`r beta_0_star_t1`) + (`r beta_1_star_t1`)x^{\prime} + (`r gamma_1_t1`)w^{\prime}$$


#### Step 2.2: Using equation 5.10, calculate alpha_2; compare to value on p. 187. Pretty close to -1, which was the transformation in Example 5.2. 

```{r}

alpha_2 <- (gamma_1_t1/beta_1_t) + alpha_1

```


$$\alpha_2 = \frac{\hat{\gamma}}{\hat{\beta_1}} + \alpha_1 =(`r alpha_2`)$$
