---
title: "Lesson 7 Lecture Example"
author: "Rick Davila"
date: "5/11/2020"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

## Lesson 7 - Install packages
#### Install necessary packages using library()
#### Perform data housekeeping - upload, name columns, display to make sure it reads properly, etc. 

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)

library(e1071)
library(xtable)
library("xlsx") # Needed to read data
library(psych) # For geometric mean in Example 5.3

rm(list = ls())

```

#### Read data file (data-ex-5-1.xlsx)

```{r}

exL <- read.xlsx("data-table-B09.xlsx",
                   sheetIndex = 1, 
                   colIndex = c(1,2,3,4,5), 
                   as.data.frame = TRUE, 
                   header = TRUE)

```

#### Assign labels to data columns using names() and attach() commands

```{r}

names(exL) <- c("fluid_vel", "viscosity","mesh_open","fluid_gas_vel","pressure_drop")
attach(exL)

```

#### Output data to make sure it reads properly

```{r results="asis"}

out <- as.data.frame(c(exL))
colnames(out) <- c("fluid_vel", "viscosity","mesh_open","fluid_gas_vel","pressure_drop")
tab <- (xtable(out, digits=c(0,2,1,2,3,1)))
print(tab, type="html")

# Output data structure and dimensions
str(exL)
dim(exL)

```

#### create multiple least squares model

```{r results="asis"}

model <- lm(pressure_drop ~ fluid_vel + viscosity + mesh_open + fluid_gas_vel)

summary(model)

xtable(summary(model))
xtable(anova(model))

```

#### make a plot of the multiple least squares model

```{r}

plot(model)

```

```{r}
# Let’s look at a plot of the R-student residuals versus the fitted values

R_Student_Residuals <- rstudent(model)
y_hat <- model$fitted.values
plot(y_hat, R_Student_Residuals, ylab = "R-Student Residuals", xlab = "Pressure Drop", main = "R-Student Residuals versus Fitted Values")


plot(fluid_vel, R_Student_Residuals, ylab = "R-Student Residuals", xlab = "Fluid Velocity", main = "R-Student Residuals versus Fluid Velocity")
plot(viscosity, R_Student_Residuals, ylab = "R-Student Residuals", xlab = "Viscosity", main = "R-Student Residuals versus Viscosity")
plot(mesh_open, R_Student_Residuals, ylab = "R-Student Residuals", xlab = "Mesh Opening", main = "R-Student Residuals versus Mesh Opening")
plot(fluid_gas_vel, R_Student_Residuals, ylab = "R-Student Residuals", xlab = "Fluid Gas Velocity", main = "R-Student Residuals versus Fluid Gas Velocity")



```

### Since there's a huge gap in the residuals versus viscosity data, we'll try a natural log on the regressor.  Redo the exercise above, but, with the lm() using log of vicosity.

```{r results="asis"}

model <- lm(pressure_drop ~ fluid_vel + log(viscosity) + mesh_open + fluid_gas_vel)

summary(model)

xtable(summary(model))
xtable(anova(model))

plot(model)

R_Student_Residuals <- rstudent(model)
y_hat <- model$fitted.values
plot(y_hat, R_Student_Residuals, ylab = "R-Student Residuals", xlab = "Pressure Drop", main = "R-Student Residuals versus Fitted Values")

plot(fluid_vel, R_Student_Residuals, ylab = "R-Student Residuals", xlab = "Fluid Velocity", main = "R-Student Residuals versus Fluid Velocity")
plot(log(viscosity), R_Student_Residuals, ylab = "R-Student Residuals", xlab = "Viscosity", main = "R-Student Residuals versus log(Viscosity)")
plot(mesh_open, R_Student_Residuals, ylab = "R-Student Residuals", xlab = "Mesh Opening", main = "R-Student Residuals versus Mesh Opening")
plot(fluid_gas_vel, R_Student_Residuals, ylab = "R-Student Residuals", xlab = "Fluid Gas Velocity", main = "R-Student Residuals versus Fluid Gas Velocity")




```




