---
title: "Lesson 7 R Activity 1"
author: "Rick Davila"
date: "5/8/2020"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

## Lesson 7 - Install packages
#### Install necessary packages using library()
#### Perform data housekeeping - upload, name columns, display to make sure it reads properly, etc. 

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)

library(e1071)
library(xtable)
library("xlsx") # Needed to read data
library(psych) # For geometric mean in Example 5.3

rm(list = ls())

```

#### Read data file (data-ex-5-1.xlsx)

```{r}

ex5_1 <- read.xlsx("data-ex-5-1.xlsx",
                   sheetIndex = 1, 
                   colIndex = c(2,3), 
                   as.data.frame = TRUE, 
                   header = TRUE)

```

#### Assign labels to data columns using names() and attach() commands

```{r}

names(ex5_1) <- c("Usage", "Demand")
attach(ex5_1)

```

#### Output data to make sure it reads properly

```{r results="asis"}

out <- as.data.frame(c(ex5_1))
colnames(out) <- c("Usage", "Demand")
tab <- (xtable(out, digits=c(0,0,2)))
print(tab, type="html")

# Output data structure and dimensions
str(ex5_1)
dim(ex5_1)

```

## Example 5.1 (p. 173-175)
#### Create scatterplot of data

```{r}

plot(Usage,Demand, main = "Scatter plot of energy demand (kW) vs energy usage (kWh)")

```

#### Create linear model and display ANOVA table. Compare to least squares fit on p. 173 and ANOVA table on p. 175

```{r}

# Obtain regression estimates using lm command
model <- lm(Demand ~ Usage)
beta_0 <- model$coefficients[1]
beta_1 <- model$coefficients[2]

# show beta estimates, std. error, tvalues and p-values
xtable(model)

```

The equation for the linear regression model: 
$$\hat{y} = (`r beta_0`) + (`r beta_1`)x$$

```{r}

# ANOVA table
# summary(aov)
out <- anova(model)

xtable(out)

```

#### Create plot of R-Student values versus fitted values

```{r}

# Demand, given Usage
model <- lm(Demand ~ Usage)
beta_0 <- model$coefficients[1]
beta_1 <- model$coefficients[2]

# show beta estimates, std. error, tvalues and p-values
xtable(model)

# Calculate y_hat
y_hat <- beta_0 + beta_1*Usage

# Calculate R-Student residuals
residuals <- rstudent(model)

plot(y_hat, residuals, main='Plot of R-student values t_i versus fitted values y_hat', xlab = 'y_hat', ylab = 'Rstudent() Residuals (t_i)')

```

#### Transform and regress on y* = sqrt(y).Compare to least squares fit on p. 173

```{r}

y_star <- sqrt(Demand)

# Demand using sqrt transformation, given Usage
model.star <- lm(y_star ~ Usage)
beta_0 <- model.star$coefficients[1]
beta_1 <- model.star$coefficients[2]

# show beta estimates, std. error, tvalues and p-values
xtable(model)

# Calculate y_hat
y_hat_star <- beta_0 + beta_1*Usage

```

#### Create plot of R-Student values versus fitted values of transformed model

```{r}

# Calculate R-Student residuals
residuals <- rstudent(model.star)

plot(y_hat_star, residuals, main='Plot of R-student values t_i versus fitted values y_hat_star', xlab = 'y_hat', ylab = 'Rstudent() Residuals (t_i)')

```

## Example 5.3 (p. 184)
#### Perform Box-Cox transformation of Electric Utility data using boxcox() function.  boxcox() outputs a range of x and y values. Plot those values and compare to p. 185. Why are the plots different?

```{r}

# create least sqares model
model <- lm(Demand ~ Usage)
library(MASS) # library containg boxcox
bc <- boxcox(Demand ~ Usage)

```

The boxcox() calcultes a log-likelhood for the y-axis versus the book using an approximate Confidence Interval for λ (based on the $SS_{Res})$.

#### Find value of lambda ("y" in R's boxcox notation) using which() and max() functions

```{r}

# extract lambda
best.lambda <- bc$x[which(bc$y==max(bc$y))]

```

The best lambda is $\lambda =`r best.lambda`$

#### Transform y (Demand) using Box-Cox transformation Equation 5.1 on p. 182. Vary lambda values and compare SS_Residual to Table 5.7 on p. 185

```{r results="asis"}

# power transformation procedure

# define power transformation function

pow_trans <- function(Lambda_In, ydata) {
# function to perform eqn 5.1 in e-book

n <- length(ydata)
#y_dot <- exp(sum(log(ydata))/n)
y_dot <- geometric.mean(ydata)


  if(Lambda_In == 0){
    y.lambda <- y_dot*log(ydata)
  
  } else {
    numerator <- ydata^(Lambda_In) - 1
    denominator <- Lambda_In*y_dot^(Lambda_In-1)
    y.lambda <- numerator/denominator

  }
result <- y.lambda
}

lambdaList <- c(-2, -1, -0.5, 0, 0.125, .25, .375, .5, best.lambda, 0.625, .75, 1, 2)
desired_length <- length(lambdaList)
SS_Res_out <- rep(NA, desired_length) # initialize a list

count <- 1
for (val in lambdaList) {

  ydata_trans <- pow_trans(val,Demand)
  model <- lm(ydata_trans ~ Usage)
  SS_Res_out[count] <- anova(model)$'Sum Sq'[2]
  count = count+1
}

List1 <- data.frame(lambdaList)
List2 <- data.frame(SS_Res_out)
out <- as.data.frame(c(List1,List2))
colnames(out) <- c("Lambda","$SS_{Res}$(Lambda)")
tab <- (xtable(out, digits=c(0, 3, 4)))
print(tab, type="html")

```






```