---
title: "Lesson 4 R Activity"
author: "Rick Davila"
date: "4/28/2020"
output:
  html_document:
    df_print: paged
  pdf_document: default
---
## Lesson 4 - Install packages

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)

library(e1071)
library(xtable)
library("xlsx") # Needed to read data
library(car) # Needed for alternative scatterplot matrix to default
library(scatterplot3d) # Needed for 3D scatterplot
library(matlib) # Needed for Invers() function
library(MASS) # Needed for ginv() function
library(standardize) # Needed for unit normal scaling in Example 3.14

rm(list = ls())

```
## Lesson 4 - Read data file (data-ex-3-1.xlsx)

```{r}

ex3_1 <- read.xlsx("data-ex-3-1.xlsx",
                   sheetIndex = 1, 
                   colIndex = c(2,3,4), 
                   as.data.frame = TRUE, 
                   header = TRUE)

```
## Lesson 4 - Assign labels to data columns using names() and attach() commands

```{r}

names(ex3_1) <- c("Delivery_Time", "Num_Cases", "Distance")
attach(ex3_1)

```
## Lesson 4 - Output data to make sure it reads properly

```{r}

# Output data to make sure it reads properly
xtable(ex3_1)

```
## Lesson 4 - Output data structure and dimensions

```{r}

# output dataframe structure
str(ex3_1)

# dim of data 'matrix' 
dim(ex3_1)

```
## Lesson 4 - Example 3.1 (p. 75-77)

#### Ex 3.1.  Create pairwise scatterplots using pairs() command
```{r}

# pairs() - The pairs function returns a plot matrix, consisting of scatterplots for each variable-combination of a data frame.

pairs(ex3_1)

```
#### Ex 3.1.  Create pairwise scatterplots using the scatterplotMatrix() command from the "car" package

```{r}

# The scatterplotMatrix function provides a convenient interface to the pairs function to produce enhanced scatterplot matrices, including univariate displays on the diagonal and a variety of fitted lines, smoothers, variance functions, and concentration ellipsoids.

scatterplotMatrix(ex3_1, use = c("pairwise.complete.obs"))

```
#### Ex 3.1.  Create 3D scatterplot using scatterplot3d() command from "scatterplot3d" package

```{r}

# Plots a three dimensional (3D) point cloud
scatterplot3d(Num_Cases, Distance, Delivery_Time)

```
#### Ex 3.1.  Obtain regression estimators using matrix algebra ("by hand")
#### Define X matrix of regressor observations (don't forget to include column for intercept)

```{r}

X <- cbind(matrix(1,length(Distance),1),as.matrix(Num_Cases),as.matrix(Distance))
y <- as.matrix(Delivery_Time)

```
#### Ex 3.1.  Display X to make sure it is correct; compare to p. 75)

```{r}
# display X matrix
X

```
#### Ex 3.1.  Define the matrix product of X_Transpose and X and display output to make sure it is correct (compare to p. 76)

```{r}

# X'X matrix
xTx <- t(X) %*% X
xTx

```

#### Ex 3.1.  Take the inverse of xTx ("x-Transpose time x") using three different approaches: Inverse(), ginv(), and inv() and display output to make sure it is correct (compare to p. 77)

```{r}

print("Inverse(xTx)")
Inverse(xTx)

print("ginv(xTx)")
ginv(xTx)

print("inv(xTx)")
inv(xTx)

```
#### Ex 3.1.  Illustration - the above matrix inverse calculations are possibly bad. How to tell? Take the inverse of the inverse using each approach and see...

```{r}

print("Inverse(Inverse(xTx))")
Inverse(Inverse(xTx))

print("ginv(ginv(xTx))")
ginv(ginv(xTx))

print("inv(inv(xTx))")
inv(inv(xTx))

```
 
 ginv(ginv(xTx)) returns the original matrix
 
#### Ex 3.1.  Continuing on... use ginv() for matrix inverse calculations from here on out.  Define / calculate X-transpose * y, where y is the vector of dependent variable observations. Compare to p. 77

```{r}

print("X-transpose * y")
t(X) %*% y

```
#### Ex 3.1.  Calculate Beta coefficients. Compare to values on p. 77.

```{r}

# The least-squares estimator of beta_coeffs

print("Beta Coefficients")
beta_hat <- ginv(xTx) %*% t(X) %*% y
beta_hat

```

So we have ˆy = 2.3412311 + 1.6159072x1 + 0.0143848x2

The equation is 
$$ \hat{y} = (`r beta_hat[1,1]`) + (`r beta_hat[2,1]`)x_1 + (`r beta_hat[3,1]`)x_2 $$
#### Ex 3.1.  Obtain regression estimates using lm command

```{r}

model <- lm(Delivery_Time ~ Num_Cases + Distance)
xtable(model)

# aov table
xtable(summary(aov(model)))

# R coefficient
res <- data.frame(model$residuals)
colnames(res) <- "Residuals"
out <- as.data.frame(c(sigma(model), summary(model)$r.squared, summary(model)$adj.r.squared))
names(out) <- ""
rownames(out) <- c("$$S$$", "$$R^2$$", "$$R^{2}_{adj}$$")
xtable(out, digits=6)

#xtable(res)

```
## Lesson 4 - Example 3.2 (p. 81)

#### Calculate SS_Residual (by hand) using the definition of a residual on p.80

$$ SS_{res} = \sum\limits_{i=1}^n \left(y_i - \hat{y_i}\right)^2$$

```{r}

y_hat <- beta_hat[1,1] + beta_hat[2,1]*Num_Cases + beta_hat[3,1]*Distance 
SS_res1 <- sum((Delivery_Time - y_hat)^2)
SS_res1

```

$$ SS_{res} = \sum\limits_{i=1}^n \left(y_i - \hat{y_i}\right)^2 = `r SS_res1` $$
#### Ex 3.2. Calculate SS_Residual using matrix algebra formula on p. 81.

```{r}

yprime_y <- t(Delivery_Time) %*% Delivery_Time
beta_x_y <- t(beta_hat)%*%t(X)%*% Delivery_Time
SS_res2 <- yprime_y - beta_x_y
SS_res2

```

The residual sum of squares using using the matrix algebra formula is:
$$ SS_{res} = \mathbf{y^{\prime}y - \hat{\beta^{\prime}}X^{\prime}y} = `r SS_res2` $$
#### Ex 3.2. Learning point: Different packages use different 'behind the curtain' approaches to calculate the matrix inverse.  For some reason, inv() and Inverse() use rounding that results in over half of the SS_Residual being lost.  Using ginv(), from the MASS package, gives results that match the textbook and the results obtained from the lm() command

```{r}

# The least-squares estimator of beta_coeffs using inv() and Inverse()

print("Beta Coefficients - using inv()")
beta_hat_inv <- inv(xTx) %*% t(X) %*% y

# yprime_y calculated in Ex 3.1
beta_x_y_inv <- t(beta_hat_inv)%*%t(X)%*% Delivery_Time
SS_res_inv <- yprime_y - beta_x_y_inv

print("Beta Coefficients - using Inverse()")
beta_hat_Inverse = Inverse(xTx) %*% t(X) %*% y

beta_x_y_Inverse <- t(beta_hat_Inverse)%*%t(X)%*% Delivery_Time
SS_res_Inverse <- yprime_y - beta_x_y_Inverse

```
The residual sum of squares using using the matrix algebra formula and inv() function is:
$$ SS_{res} = \mathbf{y^{\prime}y - \hat{\beta^{\prime}}X^{\prime}y} = `r SS_res_inv` $$
The residual sum of squares using using the matrix algebra formula and Inverse() function is:
$$ SS_{res} = \mathbf{y^{\prime}y - \hat{\beta^{\prime}}X^{\prime}y} = `r SS_res_Inverse` $$
#### Ex 3.2. Note also the discrepancy between using the definition of a residual and using the matrix algebra in Equation 3.16

```{r}

# Using the inv() function
y_hat <- beta_hat_inv[1,1] + beta_hat_inv[2,1]*Num_Cases + beta_hat_inv[3,1]*Distance 
SS_res_inv2 <- sum((Delivery_Time - y_hat)^2)

# Using the inv() function
y_hat <- beta_hat_Inverse[1,1] + beta_hat_Inverse[2,1]*Num_Cases + beta_hat_Inverse[3,1]*Distance 
SS_res_Inverse2 <- sum((Delivery_Time - y_hat)^2)

```

The residual sum of squares resulting from using inv()
$$ SS_{res} = \sum\limits_{i=1}^n \left(y_i - \hat{y_i}\right)^2 = `r SS_res_inv2` $$
The residual sum of squares resulting from using Inverse()
$$ SS_{res} = \sum\limits_{i=1}^n \left(y_i - \hat{y_i}\right)^2 = `r SS_res_Inverse2` $$
#### Ex 3.2. Calculate SST and SS_Regression (by hand). Compare to ANOVA table on p. 78. Values match.

```{r}

SS_T <- t(Delivery_Time)%*%Delivery_Time-(sum(Delivery_Time))^2/length(Delivery_Time)
SS_R <- t(beta_hat)%*%t(X)%*%Delivery_Time-(sum(Delivery_Time))^2/length(Delivery_Time)

```
$$ SS_{T} = \mathbf{y^{\prime}y}-\frac{\left(\sum\limits_{i=1}^n y_i\right)^2}{n} = `r SS_T` $$
$$ SS_{R} = \mathbf{\hat{\beta}^{\prime}X^{\prime}y}-\frac{\left(\sum\limits_{i=1}^n y_i\right)^2}{n} = `r SS_R` $$
The $SS_T$ and $SS_R$ calculated values are consistent with the values reported in the book's ANOVA table ($5784.5$ for $SS_T$ and $5550.8$ for $SS_R$).

#### Ex 3.2. Calculate ANOVA table using anova() and display the output

```{r}

# anova table, model calculated using lm() in Ex 3.1 abov
anova(model)
summary(anova(model))

```
## Lesson 4 - Example 3.3 (p. 87)

#### Test for significance of regression using F-test (by hand)

#### Calculate degrees of freedom. Output to make sure it is correct; compare to the ANOVA table on p. 87. The test for significance is a test to determine if there is a linear relationship between the response and any of the regressor variables.

```{r}

# degrees of freedom
k <- 2                     # number of beta regressor parameters
n <- length(Delivery_Time) # number of obervations

SS_reg_df <- k
SS_res_df <- n-k-1
SS_T_df <- n - 1

```


$$ SS_{R}df = `r SS_reg_df` $$
$$ SS_{res}df = `r SS_res_df` $$
$$ SS_{T}df = `r SS_T_df`$$
#### Ex 3.3.  Calculate Mean-Square for Regression and Residual

```{r}

# n = length(Delivery_Time); calculated in previous chunk

SS_T <- t(Delivery_Time)%*%Delivery_Time - (sum(Delivery_Time))^2/n
SS_R <- t(beta_hat) %*% t(X) %*% Delivery_Time - (sum(Delivery_Time))^2/n
SS_res <- SS_T - SS_R

```
$$ SS_{R} = `r SS_R` $$
$$ SS_{res} = `r SS_res` $$
#### Ex 3.3.  Calculate F-test algebraically and F-critical using qf() command

```{r}

# recall, k = 2 (above)

MS_R <- SS_R/k
MS_res <- SS_res/(n-k-1)
F_0 <- MS_R/MS_res

# F critical using the qf() command at the 0.01 significance
siglevel <- 0.01

Fcritical <- qf(1-siglevel, SS_reg_df, SS_res_df)

```

The $F_0$ statistic is $$F_0 = \frac{MS_R}{MS_{res}}=\frac{`r MS_R`}{`r MS_res`}=`r F_0` $$

The $F$ critical value is $$ F_0 = `r Fcritical` $$

## Lesson 4 - Example 3.4 (p. 88-89)

#### Test the significance of the individual regression coefficients (by hand)
#### Define C matrix for use in computing se(B_j)

```{r}

# xTx matrix calculated in the begining of this lesson R activity
C_matrix <- ginv(xTx)
C_matrix

```
#### Ex 3.4. (p. 88-89) Compute the t-test algebraically and t-critical using qt() command. Calculate p-value using pt() command.

```{r}

# n is the number of observations and p is the number of beta parameters
p <- length(beta_hat)
sigma_hat_sq = SS_res/(n - p)

# calculate t statistic for regressor parameters beta_1 and beta_2
t0_beta2 = beta_hat[2,1]/sqrt(sigma_hat_sq * C_matrix[2,2])
t0_beta3 = beta_hat[3,1]/sqrt(sigma_hat_sq * C_matrix[3,3])

# t-critical and p-value calculations
# for alpha = 0.05
alpha = 0.05

tcritical_value = abs(qt(alpha/2, df = SS_res_df))
p_value_beta2 = 2*pt(-abs(t0_beta2), df = SS_res_df)
p_value_beta3 = 2*pt(-abs(t0_beta3), df = SS_res_df)

```

The test statistics for $\beta_1$ and $\beta_2$ are $`r t0_beta2`$ and $`r t0_beta2`$, respectively.  The t critical value is $$ t_{0.025,22} = `r tcritical_value` $$ so we conclude that each regressor individually, number of cases and distance, contribute significantly to the model.  The P values are $$P(\beta1) = `r p_value_beta2`$$ and $$P(\beta2) = `r p_value_beta3`$$

## Lesson 4 - Example 3.5 (p. 92-93)

#### Perform partial F-test on the significance of the contribution of Distance to the full model
#### Ex 3.5. Create full model using lm() command

```{r}

y_bar = sum(Delivery_Time)/length(Delivery_Time)
model_full <- lm(Delivery_Time ~ Num_Cases + Distance)

beta_0 <- model_full$coefficients[1]
beta_1 <- model_full$coefficients[2]
beta_2 <- model_full$coefficients[3]

y_hat <- beta_0 + beta_1*Num_Cases + beta_2*Distance 
SS_R_mf <- sum((y_hat - y_bar)^2) 

```
#### Ex 3.5. Create reduced model using lm() command

```{r}
# excludes the Distance parameter

model_reduced <- lm(Delivery_Time ~ Num_Cases)
beta_0 <- model_reduced$coefficients[1]
beta_1 <- model_reduced$coefficients[2]

y_hat <- beta_0 + beta_1*Num_Cases
SS_R_mr <- sum((y_hat - y_bar)^2)

```
#### Ex 3.5. Create F-test algebraically and F-critical using qf() command

```{r}

r <- length(model_full$coefficients) - length(model_reduced$coefficients)
p <- length(model_full$coefficients)
n <- length(Delivery_Time)

SS_R_beta2contrib = SS_R_mf - SS_R_mr

F_0 <- (SS_R_beta2contrib/r)/MS_res

alpha = 0.05
F_critical <- qf(1-alpha,r,n-p)

```

$$F_0=\frac{\beta_2|\beta_1,\beta_0/1}{MS_{Res}}=\frac{`r SS_R_beta2contrib`}{`r MS_res`}=`r F_0`$$
Since $F_{0.05,1,22} = `r F_critical`$, we conclude that Distance (x2) contributes significantly to the model.

#### Ex 3.5. Faster way with R - Use anova() command

```{r}

# use anova() function on the ful model
F_0anova <- anova(model_full)$F[2]

```
The $F_0$ test statistic from running anova on the full model is 
$$F_0 \gt `r F_0anova`$$