---
title: "L15Ex_JetTurbine_Rick_Davila"
author: "Rick Davila"
date: "6/05/2020"
output:
  pdf_document: default
  md_document:
    variant: markdown_github
  html_document:
    df_print: paged
---

#### comment out -- use in the case where "html" is used
#### ```{r setup, include=TRUE, results="asis"}

#### Perform data housekeeping - upload, name columns, display to make sure it reads properly, etc. 
```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)

#Sys.setenv(JAVA_HOME='C:\\Program Files\\Java\\jdk-14.0.1') # for 64-bit version
#library(rJava)

library("xlsx") # Needed to read data
library(MuMIn)
library(MASS)   # Needed for ginv() function


jt_data <- read.xlsx("data-table-B13.xlsx", sheetIndex = 1, sheetName=NULL, rowIndex=NULL, startRow=NULL, endRow=NULL, colIndex= NULL, as.data.frame=TRUE, header=TRUE, colClasses=NA, keepFormulas=FALSE, encoding="unknown")

# Give labels to data columns
names(jt_data) <- c("thrust", "primary", "secondary", "fuel", "press", "exhaust", "ambient")
attach(jt_data)

# Output data to make sure it reads properly
jt_data # Note: On the spreadsheet provided on the course website, Observation #7 is 580 but in Table 8.1 and subsequent textbook results tables it is 680.

# Output data structure and dimensions
str(jt_data)
dim(jt_data)

jt_lm <- lm(thrust~primary+secondary+fuel+press+exhaust+ambient, data=jt_data, na.action = "na.fail") # Linear model of raw data

# Use dredge() function to automatically perform all regressors regression
combinations <- dredge(jt_lm, extra = c(R_Sq = function(x) summary(x)$r.squared,R_Sq_Adj = function(x) summary(x)$adj.r.squared, MS_Res = function(x) summary(x)$sigma^2,Cp, MallowCp = function(x) summary(x)$sigma^2*df.residual(x)/summary(jt_lm)$sigma^2-dim(jt_data)[1]+2*length(x$coefficients)))
print(combinations)

# Note - the book calculates Cp very differently than R does. The "Cp" column is done automatically through R and the 'MallowCp' column calculates it manually using the formula in the book.

# Reproduce the analysis for the candidate models in the video.


```

#### Influence analysis from the full model

```{r}

# full model lm()
model <- lm(thrust~primary+secondary+fuel+press+exhaust+ambient) 
summary(model)
anova(model)

# sequence of observations
Obs <- seq(1, length(thrust))

influence_stats <- data.frame(cbind(Obs))

# R-student
r_student <- rstudent(model)
influence_stats$r_student <- data.frame(cbind(r_student))

# calculate hat matrix automatically
hat_diags <- lm.influence(model)$hat
influence_stats$hat_ii <- data.frame(cbind(hat_diags))

#### this section as a check on observation 14
X <- cbind(matrix(1,length(thrust),1),
           as.matrix(primary),
           as.matrix(secondary),
           as.matrix(fuel),
           as.matrix(press),
           as.matrix(exhaust),
           as.matrix(ambient))
y <- as.matrix(thrust)

xTx <- t(X) %*% X
H_matrix <- X %*% ginv(xTx, tol=.Machine$double.eps) %*% t(X)

# get the diagonal
diag(H_matrix)
#### this section as a check on observation 14

# Cooks D
D_i_auto <- cooks.distance(model)
influence_stats$Cooks_D <- c(D_i_auto)

#### as a check on observation 14
# Calculate studentized residuals, r_i (eqn 4.8)
e_i <- model$residuals
MS_Res <- anova(model)$'Mean Sq'[8]
r_i <- e_i/sqrt(MS_Res * (1-hat_diags))

p <- sum(hat_diags)

D_i <- ((r_i)^2/p) * (hat_diags/(1-hat_diags))

# Calculate DFFITS and DFBETAS
influence_stats$DFFITS <- c(dffits(model))
dfbetas.col <- dfbetas(model)
influence_stats$DFBETAS_0 <- c(dfbetas.col[,1])
influence_stats$DFBETAS_1 <- c(dfbetas.col[,2])
influence_stats$DFBETAS_2 <- c(dfbetas.col[,3])
influence_stats$DFBETAS_3 <- c(dfbetas.col[,4])
influence_stats$DFBETAS_4 <- c(dfbetas.col[,5])
influence_stats$DFBETAS_5 <- c(dfbetas.col[,6])
influence_stats$DFBETAS_6 <- c(dfbetas.col[,7])

# Calculate Covariance Ratio
influence_stats$covratio <- c(covratio(model))

```

#### print out table

```{r results="asis"}

library(e1071)
library(xtable)

out <- influence_stats
colnames(out) <- c("Obs $i$",
                   "$rStudent$",
                   "$h_{ii}$",
                   "$D_i$",
                   "$DFFITS_i$",
                   "$DFBETAS_{0i}$",
                   "$DFBETAS_{1i}$",
                   "$DFBETAS_{2i}$",
                   "$DFBETAS_{3i}$",
                   "$DFBETAS_{4i}$",
                   "$DFBETAS_{5i}$",
                   "$DFBETAS_{6i}$",
                   "$COVRATIO_i$")

tab <- (xtable(out, digits=c(0,0,5,5,5,4,4,4,4,4,4,4,4,4)))
print(tab, type="html")

```

# Influence thresholds

```{r}

# Identify observations that exceed limits of 1 +/- 3p/n for COVRATIO
n <- length(thrust)
limit_plus <- (1 + 3*p/n)
limit_minus <- (1 - 3*p/n)
points <- which(influence_stats$covratio > limit_plus | influence_stats$covratio < limit_minus)

# hat matrix cutoff
hm_cutoff <- 2*p/n
pts_hm_cutoff <- which(influence_stats$hat_ii > hm_cutoff)

# Cook's cutoff
Di_cutoff <- 1.0
pts_Di_cutoff <- which(influence_stats$Cooks_D > Di_cutoff)

# DFFITS cutoff
DFFITS_cutoff <- 2 * sqrt(p/n)
pts_DFFITS_cutoff <- which(abs(influence_stats$DFFITS) > DFFITS_cutoff)

# DFBETAS cutoff
DFBETAS_cutoff <- 2/(sqrt(n))

# loop through subset of matrix containing DFBETAS values
count <- 0
pts_DFBETAS_cutoff <- vector() # initialize an empty vector
for (row in 1:n) {
  for (col in 6:12) {
    
    if (abs(out[row,col]) > DFBETAS_cutoff) {
        count <- count + 1
        pts_DFBETAS_cutoff[count] <- row    # get obs from out matrix
      }    
  }
}

if (count > 0) {
  pts_DFBETAS <- unique(pts_DFBETAS_cutoff)
  }

```


Flagged Observations

(1) Point(s) $`r points`$ exceed the cutoff $COVRATIO_i$ limits of $1-\frac{3p}{n}=`r limit_minus`$ and $1+\frac{3p}{n}=`r limit_plus`$.  

(2) Point(s) $`r pts_hm_cutoff`$ exceed $h_{ii}$ cutoff of $\frac{2p}{n}=`r hm_cutoff`$.  

(3) Point(s) $`r pts_Di_cutoff`$ exceed $D_i$ cutoff of $`r Di_cutoff`$.  

(4) Point(s) $`r pts_DFFITS_cutoff`$ where $|DFFITS_i|$ exceed cutoff of $2\sqrt{\frac{p}{n}}=`r DFFITS_cutoff`$.  

(5) Point(s) $`r pts_DFBETAS`$ where $|DFBETAS_{j,i}|$ exceed cutoff of $\frac{2}{\sqrt{n}} = `r DFBETAS_cutoff`$.  

This is consistent with the video where "problem children" observations 11, 20 and 28 are flagged.

#### Model A: per video

```{r}

library(car) # Needed for vif() function

# linear regression model - Model A
model.A <- lm(thrust~primary+fuel+exhaust+ambient) 
vif(model.A)
summary(model.A)
anova(model.A)

pr <- resid(model.A)/(1 - lm.influence(model.A)$hat)
press_stat <- sum(pr^2)

# Press Statistic
press_stat

```

Model A Results:
There's a relationship (overall F and p values )
Calculated PRESS Statistic: $`r press_stat`$
Cp: 5.6 (video)

VIF's -- all above 5 -- collinarity is present

#### Perform full analysis on Model A

```{r}

qqnorm(rstudent(model.A),main="Normal QQ plot of R-student residuals")
qqline(rstudent(model.A))

plot(model.A$fitted.values,rstudent(model.A), main="r-student residuals vs fits (thrust)",
     ylab="r-student residual",
     xlab="fitted value")
abline(0, 0, col="gray")

Obs <- seq(1, length(thrust))
plot(Obs, rstudent(model.A),  main = "Observations versus Order", xlab = "Observation Order", ylab = "r-student residual")
lines(Obs, rstudent(model.A))
abline(0, 0, col="gray")

plot(primary,rstudent(model.A), main="r-student residuals vs primary",
     ylab="r-student residual",
     xlab="primary")
abline(0, 0, col="gray")

plot(fuel,rstudent(model.A), main="r-student residuals vs fuel", 
     ylab="r-student residual",
     xlab="fuel")
abline(0, 0, col="gray")

plot(exhaust,rstudent(model.A), main="r-student residuals vs exhaust",
     ylab="r-student residual",
     xlab="exhaust")
abline(0, 0, col="gray")

plot(ambient,rstudent(model.A), main="r-student residuals vs ambient",
     ylab="r-student residual",
     xlab="ambient")
abline(0, 0, col="gray")

```

Model A
Normality Plot vs residuals -- no issues
Residuals vs fits -- random scatter vs fits -- no issues
Residuals vs Order -- no major issues
Residuals vs primary, fuel, exhaust, ambient -- good random scatter

#### Influence analysis of model A

```{r}

# sequence of observations
Obs <- seq(1, length(thrust))

influence_stats <- data.frame(cbind(Obs))

# R-student
r_student <- rstudent(model.A)
influence_stats$r_student <- data.frame(cbind(r_student))

# calculate hat matrix automatically
hat_diags <- lm.influence(model.A)$hat
influence_stats$hat_ii <- data.frame(cbind(hat_diags))

#### this section as a check on observation 14
X <- cbind(matrix(1,length(thrust),1),
           as.matrix(primary),
           as.matrix(fuel),
           as.matrix(exhaust),
           as.matrix(ambient))
y <- as.matrix(thrust)

xTx <- t(X) %*% X
H_matrix <- X %*% ginv(xTx, tol=.Machine$double.eps) %*% t(X)

# get the diagonal
diag(H_matrix)
#### this section as a check on observation 14

# Cooks D
D_i_auto <- cooks.distance(model.A)
influence_stats$Cooks_D <- c(D_i_auto)

#### as a check on observation 14
# Calculate studentized residuals, r_i (eqn 4.8)
e_i <- model.A$residuals
MS_Res <- anova(model.A)$'Mean Sq'[8]
r_i <- e_i/sqrt(MS_Res * (1-hat_diags))

p <- sum(hat_diags)

D_i <- ((r_i)^2/p) * (hat_diags/(1-hat_diags))

# Calculate DFFITS and DFBETAS
influence_stats$DFFITS <- c(dffits(model.A))
dfbetas.col <- dfbetas(model.A)
influence_stats$DFBETAS_0 <- c(dfbetas.col[,1])
influence_stats$DFBETAS_1 <- c(dfbetas.col[,2])
influence_stats$DFBETAS_2 <- c(dfbetas.col[,3])
influence_stats$DFBETAS_3 <- c(dfbetas.col[,4])
influence_stats$DFBETAS_4 <- c(dfbetas.col[,5])

# Calculate Covariance Ratio
influence_stats$covratio <- c(covratio(model.A))

# Identify observations that exceed limits of 1 +/- 3p/n for COVRATIO
n <- length(thrust)
limit_plus <- (1 + 3*p/n)
limit_minus <- (1 - 3*p/n)
points <- which(influence_stats$covratio > limit_plus | influence_stats$covratio < limit_minus)

# hat matrix cutoff
hm_cutoff <- 2*p/n
pts_hm_cutoff <- which(influence_stats$hat_ii > hm_cutoff)

# Cook's cutoff
Di_cutoff <- 1.0
pts_Di_cutoff <- which(influence_stats$Cooks_D > Di_cutoff)

# DFFITS cutoff
DFFITS_cutoff <- 2 * sqrt(p/n)
pts_DFFITS_cutoff <- which(abs(influence_stats$DFFITS) > DFFITS_cutoff)

# DFBETAS cutoff
DFBETAS_cutoff <- 2/(sqrt(n))

# loop through subset of matrix containing DFBETAS values
count <- 0
pts_DFBETAS_cutoff <- vector() # initialize an empty vector
for (row in 1:n) {
  for (col in 6:10) {
    
    if (abs(out[row,col]) > DFBETAS_cutoff) {
        count <- count + 1
        pts_DFBETAS_cutoff[count] <- row    # get obs from out matrix
      }    
  }
}

if (count > 0) {
  pts_DFBETAS <- unique(pts_DFBETAS_cutoff)
  }

```

Flagged Observations

(1) Point(s) $`r points`$ exceed the cutoff $COVRATIO_i$ limits of $1-\frac{3p}{n}=`r limit_minus`$ and $1+\frac{3p}{n}=`r limit_plus`$.  

(2) Point(s) $`r pts_hm_cutoff`$ exceed $h_{ii}$ cutoff of $\frac{2p}{n}=`r hm_cutoff`$.  

(3) Point(s) $`r pts_Di_cutoff`$ exceed $D_i$ cutoff of $`r Di_cutoff`$.  

(4) Point(s) $`r pts_DFFITS_cutoff`$ where $|DFFITS_i|$ exceed cutoff of $2\sqrt{\frac{p}{n}}=`r DFFITS_cutoff`$.  

(5) Point(s) $`r pts_DFBETAS`$ where $|DFBETAS_{j,i}|$ exceed cutoff of $\frac{2}{\sqrt{n}} = `r DFBETAS_cutoff`$.  

Summary
couple of leverage points
Point 20 drops DFFITS
Point 20 is a challenge to the model
Model A mitigates issues with points 20 and 28.

#### Model B: per video

```{r}

library(car) # Needed for vif() function

# linear regression model - Model A
model.B <- lm(thrust~primary+fuel+exhaust+ambient+press) 
vif(model.B)
summary(model.B)
anova(model.B)

pr <- resid(model.B)/(1 - lm.influence(model.B)$hat)
press_stat <- sum(pr^2)

# Press Statistic
press_stat

```

Model B initial review
5 term model -- overall significance
R^2 a little higher ... but, we are using 5 terms
Press: 34081

VIF worse than model but better than full model
might not need pressure 

trade off between rule and pressure.


#### Perform full analysis on Model B

```{r}

qqnorm(rstudent(model.B),main="Normal QQ plot of R-student residuals")
qqline(rstudent(model.B))

plot(model.B$fitted.values,rstudent(model.B), main="r-student residuals vs fits (thrust)",
     ylab="r-student residual",
     xlab="fitted value")
abline(0, 0, col="gray")

Obs <- seq(1, length(thrust))
plot(Obs, rstudent(model.B),  main = "Observations versus Order", xlab = "Observation Order", ylab = "r-student residual")
lines(Obs, rstudent(model.B))
abline(0, 0, col="gray")

plot(primary,rstudent(model.B), main="r-student residuals vs primary",
     ylab="r-student residual",
     xlab="primary")
abline(0, 0, col="gray")

plot(fuel,rstudent(model.B), main="r-student residuals vs fuel", 
     ylab="r-student residual",
     xlab="fuel")
abline(0, 0, col="gray")

plot(exhaust,rstudent(model.B), main="r-student residuals vs exhaust",
     ylab="r-student residual",
     xlab="exhaust")
abline(0, 0, col="gray")

plot(ambient,rstudent(model.B), main="r-student residuals vs ambient",
     ylab="r-student residual",
     xlab="ambient")
abline(0, 0, col="gray")

plot(press,rstudent(model.B), main="r-student residuals vs press",
     ylab="r-student residual",
     xlab="press")
abline(0, 0, col="gray")


```

Analysis summary:
minor improvements in normality and residual vs fits, regressors

#### Influence analysis of model B

```{r}

# sequence of observations
Obs <- seq(1, length(thrust))

influence_stats <- data.frame(cbind(Obs))

# R-student
r_student <- rstudent(model.B)
influence_stats$r_student <- data.frame(cbind(r_student))

# calculate hat matrix automatically
hat_diags <- lm.influence(model.B)$hat
influence_stats$hat_ii <- data.frame(cbind(hat_diags))

#### this section as a check on observation 14
X <- cbind(matrix(1,length(thrust),1),
           as.matrix(primary),
           as.matrix(fuel),
           as.matrix(exhaust),
           as.matrix(ambient),
           as.matrix(press))
y <- as.matrix(thrust)

xTx <- t(X) %*% X
H_matrix <- X %*% ginv(xTx, tol=.Machine$double.eps) %*% t(X)

# get the diagonal
diag(H_matrix)
#### this section as a check on observation 14

# Cooks D
D_i_auto <- cooks.distance(model.B)
influence_stats$Cooks_D <- c(D_i_auto)

#### as a check on observation 14
# Calculate studentized residuals, r_i (eqn 4.8)
e_i <- model.B$residuals
MS_Res <- anova(model.B)$'Mean Sq'[8]
r_i <- e_i/sqrt(MS_Res * (1-hat_diags))

p <- sum(hat_diags)

D_i <- ((r_i)^2/p) * (hat_diags/(1-hat_diags))

# Calculate DFFITS and DFBETAS
influence_stats$DFFITS <- c(dffits(model.B))
dfbetas.col <- dfbetas(model.B)
influence_stats$DFBETAS_0 <- c(dfbetas.col[,1])
influence_stats$DFBETAS_1 <- c(dfbetas.col[,2])
influence_stats$DFBETAS_2 <- c(dfbetas.col[,3])
influence_stats$DFBETAS_3 <- c(dfbetas.col[,4])
influence_stats$DFBETAS_4 <- c(dfbetas.col[,5])
influence_stats$DFBETAS_5 <- c(dfbetas.col[,6])


# Calculate Covariance Ratio
influence_stats$covratio <- c(covratio(model.B))

# Identify observations that exceed limits of 1 +/- 3p/n for COVRATIO
n <- length(thrust)
limit_plus <- (1 + 3*p/n)
limit_minus <- (1 - 3*p/n)
points <- which(influence_stats$covratio > limit_plus | influence_stats$covratio < limit_minus)

# hat matrix cutoff
hm_cutoff <- 2*p/n
pts_hm_cutoff <- which(influence_stats$hat_ii > hm_cutoff)

# Cook's cutoff
Di_cutoff <- 1.0
pts_Di_cutoff <- which(influence_stats$Cooks_D > Di_cutoff)

# DFFITS cutoff
DFFITS_cutoff <- 2 * sqrt(p/n)
pts_DFFITS_cutoff <- which(abs(influence_stats$DFFITS) > DFFITS_cutoff)

# DFBETAS cutoff
DFBETAS_cutoff <- 2/(sqrt(n))

# loop through subset of matrix containing DFBETAS values
count <- 0
pts_DFBETAS_cutoff <- vector() # initialize an empty vector
for (row in 1:n) {
  for (col in 6:11) {
    
    if (abs(out[row,col]) > DFBETAS_cutoff) {
        count <- count + 1
        pts_DFBETAS_cutoff[count] <- row    # get obs from out matrix
      }    
  }
}

if (count > 0) {
  pts_DFBETAS <- unique(pts_DFBETAS_cutoff)
  }

```

Flagged Observations

(1) Point(s) $`r points`$ exceed the cutoff $COVRATIO_i$ limits of $1-\frac{3p}{n}=`r limit_minus`$ and $1+\frac{3p}{n}=`r limit_plus`$.  

(2) Point(s) $`r pts_hm_cutoff`$ exceed $h_{ii}$ cutoff of $\frac{2p}{n}=`r hm_cutoff`$.  

(3) Point(s) $`r pts_Di_cutoff`$ exceed $D_i$ cutoff of $`r Di_cutoff`$.  

(4) Point(s) $`r pts_DFFITS_cutoff`$ where $|DFFITS_i|$ exceed cutoff of $2\sqrt{\frac{p}{n}}=`r DFFITS_cutoff`$.  

(5) Point(s) $`r pts_DFBETAS`$ where $|DFBETAS_{j,i}|$ exceed cutoff of $\frac{2}{\sqrt{n}} = `r DFBETAS_cutoff`$.  

Summary
Not as well as model A.  4 obs flagged in DFFITS.  point 11 has popped up. Model A has done a better job on influence points, however, model B has done a better job in the residual analysis


#### Model C: per video

```{r}

# linear regression model - Model C
model.C <- lm(thrust~primary+press+exhaust+ambient) 
vif(model.C)
summary(model.C)
anova(model.C)

pr <- resid(model.C)/(1 - lm.influence(model.C)$hat)
press_stat <- sum(pr^2)

# Press Statistic
press_stat

```

Model B initial review
5 term model -- overall significance
R^2 a little higher ... but, we are using 5 terms
Press: 34081

VIF worse than model but better than full model
might not need pressure 

trade off between rule and pressure.


#### Perform full analysis on Model C

```{r}

qqnorm(rstudent(model.C),main="Normal QQ plot of R-student residuals")
qqline(rstudent(model.C))

plot(model.C$fitted.values,rstudent(model.C), main="r-student residuals vs fits (thrust)",
     ylab="r-student residual",
     xlab="fitted value")
abline(0, 0, col="gray")

Obs <- seq(1, length(thrust))
plot(Obs, rstudent(model.C),  main = "Observations versus Order", xlab = "Observation Order", ylab = "r-student residual")
lines(Obs, rstudent(model.C))
abline(0, 0, col="gray")

plot(primary,rstudent(model.C), main="r-student residuals vs primary",
     ylab="r-student residual",
     xlab="primary")
abline(0, 0, col="gray")

plot(exhaust,rstudent(model.C), main="r-student residuals vs exhaust",
     ylab="r-student residual",
     xlab="exhaust")
abline(0, 0, col="gray")

plot(ambient,rstudent(model.C), main="r-student residuals vs ambient",
     ylab="r-student residual",
     xlab="ambient")
abline(0, 0, col="gray")

plot(press,rstudent(model.C), main="r-student residuals vs press",
     ylab="r-student residual",
     xlab="press")
abline(0, 0, col="gray")


```

Analysis summary:
minor improvements in normality and residual vs fits, regressors

#### Influence analysis of model C

```{r}

# sequence of observations
Obs <- seq(1, length(thrust))

influence_stats <- data.frame(cbind(Obs))

# R-student
r_student <- rstudent(model.C)
influence_stats$r_student <- data.frame(cbind(r_student))

# calculate hat matrix automatically
hat_diags <- lm.influence(model.C)$hat
influence_stats$hat_ii <- data.frame(cbind(hat_diags))

#### this section as a check on observation 14
X <- cbind(matrix(1,length(thrust),1),
           as.matrix(primary),
           as.matrix(exhaust),
           as.matrix(ambient),
           as.matrix(press))
y <- as.matrix(thrust)

xTx <- t(X) %*% X
H_matrix <- X %*% ginv(xTx, tol=.Machine$double.eps) %*% t(X)

# get the diagonal
diag(H_matrix)
#### this section as a check on observation 14

# Cooks D
D_i_auto <- cooks.distance(model.C)
influence_stats$Cooks_D <- c(D_i_auto)

#### as a check on observation 14
# Calculate studentized residuals, r_i (eqn 4.8)
e_i <- model.C$residuals
MS_Res <- anova(model.C)$'Mean Sq'[8]
r_i <- e_i/sqrt(MS_Res * (1-hat_diags))

p <- sum(hat_diags)

D_i <- ((r_i)^2/p) * (hat_diags/(1-hat_diags))

# Calculate DFFITS and DFBETAS
influence_stats$DFFITS <- c(dffits(model.C))
dfbetas.col <- dfbetas(model.C)
influence_stats$DFBETAS_0 <- c(dfbetas.col[,1])
influence_stats$DFBETAS_1 <- c(dfbetas.col[,2])
influence_stats$DFBETAS_2 <- c(dfbetas.col[,3])
influence_stats$DFBETAS_3 <- c(dfbetas.col[,4])
influence_stats$DFBETAS_4 <- c(dfbetas.col[,5])

# Calculate Covariance Ratio
influence_stats$covratio <- c(covratio(model.C))

# Identify observations that exceed limits of 1 +/- 3p/n for COVRATIO
n <- length(thrust)
limit_plus <- (1 + 3*p/n)
limit_minus <- (1 - 3*p/n)
points <- which(influence_stats$covratio > limit_plus | influence_stats$covratio < limit_minus)

# hat matrix cutoff
hm_cutoff <- 2*p/n
pts_hm_cutoff <- which(influence_stats$hat_ii > hm_cutoff)

# Cook's cutoff
Di_cutoff <- 1.0
pts_Di_cutoff <- which(influence_stats$Cooks_D > Di_cutoff)

# DFFITS cutoff
DFFITS_cutoff <- 2 * sqrt(p/n)
pts_DFFITS_cutoff <- which(abs(influence_stats$DFFITS) > DFFITS_cutoff)

# DFBETAS cutoff
DFBETAS_cutoff <- 2/(sqrt(n))

# loop through subset of matrix containing DFBETAS values
count <- 0
pts_DFBETAS_cutoff <- vector() # initialize an empty vector
for (row in 1:n) {
  for (col in 6:10) {
    
    if (abs(out[row,col]) > DFBETAS_cutoff) {
        count <- count + 1
        pts_DFBETAS_cutoff[count] <- row    # get obs from out matrix
      }    
  }
}

if (count > 0) {
  pts_DFBETAS <- unique(pts_DFBETAS_cutoff)
  }

```

Flagged Observations

(1) Point(s) $`r points`$ exceed the cutoff $COVRATIO_i$ limits of $1-\frac{3p}{n}=`r limit_minus`$ and $1+\frac{3p}{n}=`r limit_plus`$.  

(2) Point(s) $`r pts_hm_cutoff`$ exceed $h_{ii}$ cutoff of $\frac{2p}{n}=`r hm_cutoff`$.  

(3) Point(s) $`r pts_Di_cutoff`$ exceed $D_i$ cutoff of $`r Di_cutoff`$.  

(4) Point(s) $`r pts_DFFITS_cutoff`$ where $|DFFITS_i|$ exceed cutoff of $2\sqrt{\frac{p}{n}}=`r DFFITS_cutoff`$.  

(5) Point(s) $`r pts_DFBETAS`$ where $|DFBETAS_{j,i}|$ exceed cutoff of $\frac{2}{\sqrt{n}} = `r DFBETAS_cutoff`$.  

Summary
Multicollinearity exist ... as it does in all models A and B.  Normal probability plot looks O.K.  Residuals vs fits looks good.  Model C. shows most leverage points.  

```{r}

library(olsrr)

k <- ols_step_forward_p(model, penter=0.1)
plot(k)

# final model
k$model

k <- ols_step_backward_p(model, prem =0.1)
plot(k)

# final model
k$model

k <- ols_step_both_p(model, prem =0.1)
plot(k)

# final model
k$model


```

