---
title: "Lesson 9 Lecture Example"
author: "Rick Davila"
date: "5/16/2020"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

## Lesson 9 - Install packages
#### Install necessary packages using library()
#### Perform data housekeeping - upload, name columns, display to make sure it reads properly, etc. 

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)

library(e1071)
library(xtable)
library("xlsx") # Needed to read data
library(psych) # For geometric mean in Example 5.3
library(MASS) # Needed for ginv() function

rm(list = ls())

```

#### Read data file (data-table-B09.xlsx)

```{r}

exL <- read.xlsx("data-table-B09.xlsx",
                   sheetIndex = 1, 
                   colIndex = c(1,2,3,4,5), 
                   as.data.frame = TRUE, 
                   header = TRUE)

```

#### Assign labels to data columns using names() and attach() commands

```{r}

names(exL) <- c("fluid_vel", "viscosity","mesh_open","fluid_gas_vel","pressure_drop")
attach(exL)

```

#### Output data to make sure it reads properly

```{r results="asis"}

out <- as.data.frame(c(exL))
colnames(out) <- c("fluid_vel", "viscosity","mesh_open","fluid_gas_vel","pressure_drop")
tab <- (xtable(out, digits=c(0,2,1,2,3,1)))
print(tab, type="html")

# Output data structure and dimensions
str(exL)
dim(exL)

```

#### create multiple least squares model

```{r results="asis"}

model <- lm(pressure_drop ~ fluid_vel + viscosity + mesh_open + fluid_gas_vel)

summary(model)

xtable(summary(model))
xtable(anova(model))

```

#### make a plot of the multiple least squares model

```{r}

plot(model)

```

```{r}
# Let’s look at a plot of the R-student residuals versus the fitted values

R_Student_Residuals <- rstudent(model)
y_hat <- model$fitted.values
plot(y_hat, R_Student_Residuals, ylab = "R-Student Residuals", xlab = "Pressure Drop", main = "R-Student Residuals versus Fitted Values")


plot(fluid_vel, R_Student_Residuals, ylab = "R-Student Residuals", xlab = "Fluid Velocity", main = "R-Student Residuals versus Fluid Velocity")
plot(viscosity, R_Student_Residuals, ylab = "R-Student Residuals", xlab = "Viscosity", main = "R-Student Residuals versus Viscosity")
plot(mesh_open, R_Student_Residuals, ylab = "R-Student Residuals", xlab = "Mesh Opening", main = "R-Student Residuals versus Mesh Opening")
plot(fluid_gas_vel, R_Student_Residuals, ylab = "R-Student Residuals", xlab = "Fluid Gas Velocity", main = "R-Student Residuals versus Fluid Gas Velocity")



```

### Since there's a huge gap in the residuals versus viscosity data, we'll try a natural log on the regressor.  Redo the exercise above, but, with the lm() using log of vicosity.

```{r results="asis"}

model <- lm(pressure_drop ~ fluid_vel + log(viscosity) + mesh_open + fluid_gas_vel)

summary(model)

xtable(summary(model))
xtable(anova(model))

plot(model)

R_Student_Residuals <- rstudent(model)
y_hat <- model$fitted.values
plot(y_hat, R_Student_Residuals, ylab = "R-Student Residuals", xlab = "Pressure Drop", main = "R-Student Residuals versus Fitted Values")

plot(fluid_vel, R_Student_Residuals, ylab = "R-Student Residuals", xlab = "Fluid Velocity", main = "R-Student Residuals versus Fluid Velocity")
plot(log(viscosity), R_Student_Residuals, ylab = "R-Student Residuals", xlab = "Viscosity", main = "R-Student Residuals versus log(Viscosity)")
plot(mesh_open, R_Student_Residuals, ylab = "R-Student Residuals", xlab = "Mesh Opening", main = "R-Student Residuals versus Mesh Opening")
plot(fluid_gas_vel, R_Student_Residuals, ylab = "R-Student Residuals", xlab = "Fluid Gas Velocity", main = "R-Student Residuals versus Fluid Gas Velocity")

```

# Lesson 9 additions

```{r}
exL2 <- read.xlsx("data-table-B13.xlsx",
                   sheetIndex = 1, 
                   colIndex = c(2,3,4,5,6,7,8), 
                   as.data.frame = TRUE, 
                   header = TRUE)

names(exL2) <- c("thrust",
                 "primary",
                 "secondary",
                 "fuel",
                 "press",
                 "exhaust",
                 "ambient")
attach(exL2)

# Output data structure and dimensions
str(exL2)
dim(exL2)

```

```{r results="asis"}

out <- as.data.frame(c(exL2))
colnames(out) <- c("thrust", "primary","secondary","fuel","press","exhaust","ambient")
tab <- xtable(out, digits=c(0,0,0,0,0,0,0,0))
print(tab, type="html")

# Perform preliminary calculations that we might use at some point
N <- length(thrust)

X <- matrix(c(rep(1, N), 
              exL2[, "primary"],
              exL2[, "secondary"],
              exL2[, "fuel"],
              exL2[, "press"],
              exL2[, "exhaust"],
              exL2[, "ambient"]), ncol = 7, byrow = FALSE) # Define X matrix of regressor observations
xTx <- t(X) %*% X # Calulate the matrix product of X_Transpose and X
C <- ginv(xTx, tol=.Machine$double.eps) # Define C matrix


```


Jet Turbine Ex -- lm() fit

```{r}

model.jet <- lm(thrust ~ primary + secondary + fuel + press + exhaust + ambient)

model.jet
summary(model.jet)
anova(model.jet)

df_T <- N - 1
df_Regression <- length(model.jet$coefficients) - 1 # Number of predictor variables, two in this case
k <- df_Regression # Sometimes helpful to use the variable
p <- k + 1 # Same reasoning
df_Residual <- df_T - df_Regression 

y_observed <- exL2[, "thrust"] # Create vector of dependent variable observations
y_predicted <- (X %*% ginv(t(X)%*%X,tol=.Machine$double.eps) %*% t(X)) %*% y_observed # Hat matrix approach, just for grins.
y_bar <- mean(y_observed)

xTy <- t(X) %*% y_observed # Define rhs of least squares normal equations in matrix form
BetaCoefficients <- C %*% xTy # Calculate using ginv()
ResidualVector <- y_observed - X%*%BetaCoefficients # Define vector of residuals for each observation
SS_Residuals <- t(ResidualVector) %*% ResidualVector # Dot product of residual vector and its transpose

SST_dt <- sum((y_observed - y_bar)^2)

SS_Regression_dt <- sum((y_predicted - y_bar)^2)
SS_Residual_dt <- SST_dt - SS_Regression_dt

MS_Regression <- SS_Regression_dt / df_Regression
MS_Residual <- SS_Residual_dt / df_Residual

F_test <- MS_Regression / MS_Residual
t_test <- BetaCoefficients / sqrt(MS_Residual * diag(C)) # Formula on p.88

significanceLevel <- 0.05
F_critical <- qf(1-significanceLevel/2, df_Regression, df_Residual)
t_critical <- qt(1-significanceLevel/2, df_Residual)
#p_val <- 2*(1 - pt(t_test, df_Residual)) # Multiply by two since it is a two-tailed hypothesis test

# compare with my calculations
alpha = 0.05

tcritical_value = abs(qt(alpha/2, df = df_Residual))
p_values = 2*pt(-abs(t_test), df = df_Residual)



```

```{r}

qqnorm(model.jet$residuals,main="Normal QQ plot of Residuals (e_i)")
qqline(model.jet$residuals)

plot(thrust,rstudent(model.jet), main="residuals vs fits (thrust)",
     ylab="r-student residual")
abline(0, 0, col="gray")

plot(primary,rstudent(model.jet), main="residuals vs primary",
     ylab="r-student residual")
abline(0, 0, col="gray")

plot(secondary,rstudent(model.jet), main="residuals vs secondary",
     ylab="r-student residual")
abline(0, 0, col="gray")

plot(fuel,rstudent(model.jet), main="residuals vs fuel", 
     ylab="r-student residual")
abline(0, 0, col="gray")

plot(press,rstudent(model.jet), main="residuals vs press", 
     ylab="r-student residual")
abline(0, 0, col="gray")

plot(exhaust,rstudent(model.jet), main="residuals vs exhaust",
     ylab="r-student residual")
abline(0, 0, col="gray")

plot(ambient,rstudent(model.jet), main="residuals vs ambient",
     ylab="r-student residual")
abline(0, 0, col="gray")

```



