---
title: "Lesson 9 R Activity"
author: "Rick Davila"
date: "5/13/2020"
output:
  html_document:
    df_print: paged
  word_document: default
  pdf_document: default
---

## Lesson 9 - Install packages
#### Perform data housekeeping - upload, name columns, display to make sure it reads properly, etc. 

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)

library(e1071)
library(xtable)
library("xlsx") # Needed to read data
library(MASS) # Needed for ginv() function

rm(list = ls())

```

#### Upload data-ex-3-1.xlsx data file and label columns

```{r results="asis"}
exL9 <- read.xlsx("data-ex-3-1.xlsx",
                   sheetIndex = 1, 
                   colIndex = c(2,3,4), 
                   as.data.frame = TRUE, 
                   header = TRUE)

# Assign labels to data columns using names() and attach() commands
names(exL9) <- c("time","cases","distance")
attach(exL9) 

# Output data to make sure it reads properly
out <- as.data.frame(c(exL9))
colnames(out) <- c("time","cases","distance")
tab <- (xtable(out, digits=c(0,2,0,0)))
print(tab, type="html")

# Output data structure and dimensions
str(exL9)
dim(exL9)

```

## Example 6.1 (p.213-214)
#### Calculate hat matrix values (by hand)

```{r}

X <- cbind(matrix(1,length(distance),1),as.matrix(cases),as.matrix(distance))
y <- as.matrix(time)

xTx <- t(X) %*% X
H_matrix <- X %*% ginv(xTx, tol=.Machine$double.eps) %*% t(X)

# get the diagonal
diag(H_matrix)

```

#### Calculate hat matrix values automatically

```{r}

# perform multiple least squares regression
model <-  lm(time ~ cases+distance)

# calculate hat matrix automatically
hat_diags <- lm.influence(model)$hat
hat_diags

```

#### Create data frame to reproduce Table 6.1 on p. 214 - start with column for Observation and h_ii

```{r results="asis"}

# sequence of observations
Obs <- seq(1, length(time))

influence_stats <- data.frame(cbind(Obs, hat_diags))

out <- influence_stats
colnames(out) <- c("Obs $i$", "$h_{ii}$")
tab <- (xtable(out, digits=c(0,0,5)))
print(tab, type="html")

```

#### Create shell of unnamed table on p. 213

```{r results="asis"}

Run <- c("9 and 22 in",
         "9 out",
         "22 out",
         "9 and 22 out")
beta_0 <- c(" "," "," "," ")
beta_1 <- c(" "," "," "," ")
beta_2 <- c(" "," "," "," ")
MS_Res <- c(" "," "," "," ")
R_sqrd <- c(" "," "," "," ")

unnamed_table <- data.frame(cbind(Run,
                                  beta_0,
                                  beta_1,
                                  beta_2,
                                  MS_Res,
                                  R_sqrd))
out <- unnamed_table
colnames(out) <- c("Run",
                   "beta_hat_0",
                   "beta_hat_1",
                   "beta_hat_2",
                   "$MS_{Res}$",
                   "$R_2$")
tab <- (xtable(out, digits=c(0,0,0,0,0,0,0)))
print(tab, type="html")

```

#### Create models for the four scenarios in the unnamed table on p. 213
#### Note: Deletions using subset= are done sequentially. So, subset=(1:N)[-1][-2] removes the first observation and then the second of the remaining observations.

```{r}

# scenario 1, points 9 and 22 in
time.s1 <- time
cases.s1 <- cases
distance.s1 <- distance

model.s1 <-  lm(time.s1 ~ cases.s1 + distance.s1)

# scenario 2, point 9 out
time.s2 <- time[1:length(time)][-9]
cases.s2 <- cases[1:length(cases)][-9]
distance.s2 <- distance[1:length(distance)][-9]

model.s2 <-  lm(time.s2 ~ cases.s2 + distance.s2)

# scenario 3, point 22 out
time.s3 <- time[1:length(time)][-22]
cases.s3 <- cases[1:length(cases)][-22]
distance.s3 <- distance[1:length(distance)][-22]

model.s3 <-  lm(time.s3 ~ cases.s3 + distance.s3)

# scenario 4, points 9 and 22 out
time.s4 <- time[1:length(time)][-9][-21]
cases.s4 <- cases[1:length(cases)][-9][-21]
distance.s4 <- distance[1:length(distance)][-9][-21]

model.s4 <-  lm(time.s4 ~ cases.s4 + distance.s4)


```

#### Display completed (unnamed) table on bottom of p. 213

```{r results="asis"}

Run <- c("9 and 22 in",
         "9 out",
         "22 out",
         "9 and 22 out")

beta_0 <- as.data.frame(c(model.s1$coeff[1],
            model.s2$coeff[1],
            model.s3$coeff[1],
            model.s4$coeff[1]))

beta_1 <- as.data.frame(c(model.s1$coeff[2],
            model.s2$coeff[2],
            model.s3$coeff[2],
            model.s4$coeff[2]))

beta_2 <- as.data.frame(c(model.s1$coeff[3],
            model.s2$coeff[3],
            model.s3$coeff[3],
            model.s4$coeff[3]))

MS_Res <- as.data.frame(c(anova(model.s1)$'Mean Sq'[3],
            anova(model.s2)$'Mean Sq'[3],
            anova(model.s3)$'Mean Sq'[3],
            anova(model.s4)$'Mean Sq'[3]))

R_sqrd <- as.data.frame(c(summary(model.s1)$r.squared,
            summary(model.s2)$r.squared,
            summary(model.s3)$r.squared,
            summary(model.s4)$r.squared))

unnamed_table2 <- data.frame(cbind(Run,
                                  beta_0,
                                  beta_1,
                                  beta_2,
                                  MS_Res,
                                  R_sqrd))
out2 <- unnamed_table2

colnames(out2) <- c("Run",
                   "beta_hat_0",
                   "beta_hat_1",
                   "beta_hat_2",
                   "$MS_{Res}$",
                   "$R_2$")

rownames(out2) <- c("1","2","3","4")
tab2 <- (xtable(out2, digits=c(0,0,3,3,3,3,4)))
print(tab2, type="html")

```

## Example 6.2 (p.216)
#### Calculate Cook's D using Equation 6.5

```{r}

# rstudent residual calculation
model.1 <- lm(time ~ cases + distance)

# Calculate studentized residuals, r_i (eqn 4.8)
e_i <- model.1$residuals
MS_Res <- anova(model.1)$'Mean Sq'[3]
r_i <- e_i/sqrt(MS_Res * (1-hat_diags))

p <- sum(hat_diags)

D_i <- ((r_i)^2/p) * (hat_diags/(1-hat_diags))

D_i

```

#### Calculate Cook's D using cooks.distance(). Does this give the same answer as the "by hand" approach?

```{r}

D_i_auto <- cooks.distance(model.1)
D_i_auto

```

cooks.distance() matches the output from the by-hands approach.

#### Add Cook's D to the Table 6.1 dataframe

```{r}

# obtain and add Cook's D to table 6.1 dataframe
influence_stats$Cooks_D <- c(D_i_auto)

```

## Example 6.3 (p.218-219)
#### Calculate DFFITS and DFBETAS using R

```{r}

influence_stats$DFFITS <- c(dffits(model.1))
dfbetas.col <- dfbetas(model.1)
influence_stats$DFBETAS_0 <- c(dfbetas.col[,1])
influence_stats$DFBETAS_1 <- c(dfbetas.col[,2])
influence_stats$DFBETAS_2 <- c(dfbetas.col[,3])

```

#### Update Table 6.1

```{r results="asis"}
out <- influence_stats
colnames(out) <- c("Obs $i$",
                   "$h_{ii}$",
                   "$D_i$",
                   "$DFFITS_i$",
                   "$DFBETAS_{0i}$",
                   "$DFBETAS_{1i}$",
                   "$DFBETAS_{2i}$")

tab <- (xtable(out, digits=c(0,0,5,5,4,4,4,4)))
print(tab, type="html")

```

## Example 6.4 (p. 219)
#### Calculate Covariance Ratio using R

```{r}

influence_stats$covratio <- c(covratio(model.1))

```

#### Update Table 6.1

```{r results="asis"}

out <- influence_stats
colnames(out) <- c("Obs $i$",
                   "$h_{ii}$",
                   "$D_i$",
                   "$DFFITS_i$",
                   "$DFBETAS_{0i}$",
                   "$DFBETAS_{1i}$",
                   "$DFBETAS_{2i}$",
                   "$COVRATIO_i$")

tab <- (xtable(out, digits=c(0,0,5,5,4,4,4,4,4)))
print(tab, type="html")

```

#### Identify observations that exceed limits of 1 +/- 3p/n for COVRATIO using which() and the "or" logical operator (|). Are these the same points identified in the textbook?

```{r}

n <- length(time)
limit_plus <- (1 + 3*p/n)
limit_minus <- (1 - 3*p/n)
points <- which(influence_stats$covratio > limit_plus | influence_stats$covratio < limit_minus)

```

Points $`r points`$ exceed the cutoff $COVRATIO_i$ limits of $`r limit_minus`$ and $`r limit_plus`$.  The textbook identified points 9 and 22, but not point 16.  For my calculations, point 16 barely exceeds the $`r limit_plus`$ limit.
