---
title: "Lesson 6 R Activity"
author: "Rick Davila"
date: "5/4/2020"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

## Lesson 6 - Install packages
#### Install necessary packages using library()
#### Perform data housekeeping - upload, name columns, display to make sure it reads properly, etc. 

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)

library(e1071)
library(xtable)
library("xlsx") # Needed to read data
library(car) # Needed for alternative scatterplot matrix to default
library(scatterplot3d) # Needed for 3D scatterplot
library(matlib) # Needed for Invers() function
library(MASS) # Needed for ginv() function
library(standardize) # Needed for unit normal scaling in Example 3.14

rm(list = ls())

```

#### Read data file (data-ex-3-1.xlsx)

```{r}

ex3_1 <- read.xlsx("data-ex-3-1.xlsx",
                   sheetIndex = 1, 
                   colIndex = c(2,3,4), 
                   as.data.frame = TRUE, 
                   header = TRUE)

```

#### Assign labels to data columns using names() and attach() commands

```{r}

names(ex3_1) <- c("Delivery_Time", "Num_Cases", "Distance")
attach(ex3_1)

```

#### Output data to make sure it reads properly

```{r results="asis"}

out <- as.data.frame(c(ex3_1))
colnames(out) <- c("Delivery Time, $y$ (min)", "Number of Cases, $x_i$","Distance, $x_2$ (ft)")
tab <- (xtable(out, digits=c(0,2,0,0)))
print(tab, type="html")

# Output data structure and dimensions
str(ex3_1)
dim(ex3_1)

```

## Lesson 6 - Example 4.1 (p.135-136)
#### Calculate residuals and compare to Column (1) in Table 4.1, p. 137. 

```{r}

# Define X matrix of regressor observations
X <- cbind(matrix(1,length(Distance),1),as.matrix(Num_Cases),as.matrix(Distance))
y <- as.matrix(Delivery_Time)

# X'X matrix
xTx <- t(X) %*% X

# Calculate least-squares estimator of beta_coeffs
beta_hat <- ginv(xTx) %*% t(X) %*% y
beta_0 <- beta_hat[1,1]
beta_1 <- beta_hat[2,1]
beta_2 <- beta_hat[3,1]

```

The least squares fit is $$\hat{y} = (`r beta_0`) + (`r beta_1`)x_1 + (`r beta_2`)x_2 $$

```{r results="asis"}

# least squares fit equation
y_hat <- beta_0 + beta_1*Num_Cases + beta_2*Distance
e_i=data.frame(Delivery_Time - y_hat) # e_i are the residuals
names(e_i) <- c("Res")
attach(e_i)

out <- as.data.frame(c(e_i))
colnames(out) <- c("Residuals")
tab <- (xtable(out, digits=c(0,6)))
print(tab, type="html")

```

#### Calculate standardized residuals using Equation 4.2;compare to Column (2) in Table 4.1, p. 137. 

```{r results="asis"}

n <- length(Res)       # number of residuals
ei_bar <- sum(Res)/n   # avg of residuals
k <- 2                 # number of beta regressor coefficients
p <- k + 1             # number of parameters

MS_Res <- sum((Res - ei_bar)^2)/(n-p)

# standardized residuals (eqn 4.2)
d_i <- data.frame(Res/(sqrt(MS_Res)))
names(d_i) <- c("Std_Res")
attach(d_i)

out <- as.data.frame(c(d_i))
colnames(out) <- c("Standardized Residuals")
tab <- (xtable(out, digits=c(0,6)))
print(tab, type="html")

```

#### Calculate studentized residuals using Equation 4.8; compare to Column (3) in Table 4.1, p. 137. 

```{r results="asis"}

# calculate the hat matrix
H <- X %*% ginv(xTx) %*% t(X) 

# to get the diagonal elements of H, use diag(H) function

# studentized residuals (eqn 4.8)
r_i <- data.frame(Res/(sqrt(MS_Res*(1-diag(H)))))
names(r_i) <- c("Stud_Res")
attach(r_i)

out <- as.data.frame(c(r_i))
colnames(out) <- c("Studentized Residuals")
tab <- (xtable(out, digits=c(0,6)))
print(tab, type="html")

```

#### Calculate PRESS residuals using Equation 4.11; compare to Column (5) in Table 4.1, p. 137. 

```{r results="asis"}

# PRESS residuals (eqn 4.11)
e_pip <- data.frame(Res/(1-diag(H)))
names(e_pip) <- c("PRESS_Res")
attach(e_pip)

out <- as.data.frame(c(e_pip))
colnames(out) <- c("PRESS residuals")
tab <- (xtable(out, digits=c(0,6)))
print(tab, type="html")

```

#### Calculate R-Student residuals using Equations 4.12 and 4.13; compare to Column (6) in Table 4.1, p. 137. Looks good.

```{r results="asis"}

# using eqn 4.12
Si_sqr = ((n-p)*MS_Res-(Res^2/(1-diag(H))))/(n-p-1)

# R-student (eqn 4.13)
t_i <- data.frame(Res/(sqrt(Si_sqr*(1-diag(H)))))
names(t_i) <- c("R_Stud")
attach(t_i)

out <- as.data.frame(c(t_i))
colnames(out) <- c("R-Student Residuals")
tab <- (xtable(out, digits=c(0,6)))
print(tab, type="html")

```

#### Reproduce Table 4.1 on p. 137 - Consolidate above vectors into a dataframe using data.frame(). Give column names using colnames() and c().

```{r results="asis"}

out <- as.data.frame(c(e_i, d_i, r_i, e_pip, t_i))
colnames(out) <- c(" Residuals, $e_i$ "," Standardized Residuals, $d_i$ "," Studentized Residuals, $r_i$ "," PRESS Residuals, $e_{(i)}$ "," R-student, $t_i$ ")
tab <- (xtable(out, digits=c(0, 6, 6, 6, 6, 6)))
print(tab, type="html")

```

#### Note: R has functions called rstandard() and rstudent(). Note that rstandard() outputs what the textbook calls Studentized Residuals and rstudent() outputs what the textbook calls R-Student residuals or 'externally studentized' residuals. Don't get confused.  Compare your manually-computer vectors above with the rstandard() and rstudent() functions. The rounded difference should be a vector of zeros. 

```{r results="asis"}

# test rstandard() function
test.lm <- lm(Delivery_Time~ Num_Cases + Distance)
test.rstandard <- data.frame(rstandard(test.lm))
test.rstudent <- data.frame(rstudent(test.lm))

out <- as.data.frame(c(e_i, d_i, r_i, e_pip, t_i, test.rstandard, test.rstudent))
colnames(out) <- c(" Residuals, $e_i$ "," Standardized Residuals, $d_i$ "," Studentized Residuals, $r_i$ "," PRESS Residuals, $e_{(i)}$ "," R-student, $t_i$ ", " rstandard() ", " rstudent() ")
tab <- (xtable(out, digits=c(0, 6, 6, 6, 6, 6, 6, 6)))
print(tab, type="html")

```

## Lesson 6 - Example 4.2 (p. 139-140)
#### Create plot of residuals against fitted y_hat values for each type of residual calculated in Example 4.1 using plot()

```{r}

# before plotting, get colors
cols <- palette()

# plot the first curve by calling plot() function
plot(y_hat, Res, ylab = "Residuals", xlab = "Fitted Value", col=cols[1], pch=1, ylim=range(Res, Std_Res, Stud_Res, PRESS_Res))
points(y_hat, Std_Res, col=cols[2], pch=2)
points(y_hat, Stud_Res, col=cols[3], pch=3)
points(y_hat, PRESS_Res, col=cols[4], pch=4)
points(y_hat, R_Stud, col=cols[5], pch=5)

abline(h=0, col="gray")

legend("top",legend=c("Residuals","Standardized","Studentized","PRESS","R-Student"), col=c(cols[1],cols[2],cols[3],cols[4],cols[5]), pch=c(1, 2, 3, 4, 5), ncol = 3)

```

#### Create normal probability plot of residuals using qqnorm()

```{r}

# the normal probability plot is a graphical technique for assessing whether or not a data set is approximately normally distributed.  Departures from a straight line fit indicates depature from normality

qqnorm(Res,main="Normal QQ plot of Residuals (e_i)")
qqline(Res)

qqnorm(Std_Res,main="Normal QQ plot of Standardized Residuals")
qqline(Std_Res)

qqnorm(Stud_Res,main="Normal QQ plot of Studentized Residuals")
qqline(Stud_Res)

qqnorm(PRESS_Res,main="Normal QQ plot of PRESS Residuals")
qqline(PRESS_Res)

qqnorm(R_Stud,main="Normal QQ plot of R-Student Residuals")
qqline(R_Stud)

```

## Lesson 6 - Example 4.3 (p. 141)
#### Create a plot of residuals against the each of the regressors (Use R-Student residuals)

```{r}

# plot number of cases versus R-student residuals
plot(Num_Cases, R_Stud, ylab = "R-Student Residuals", xlab = "Number of Cases", main = "Residuals vs Number of Cases", pch=16)
abline(h=0, col="gray")

# plot Distance versus R-student residuals
plot(Distance, R_Stud, ylab = "R-Student Residuals", xlab = "Distance (ft)", main = "Residuals vs Distance", pch=16)
abline(h=0, col="gray")

```

## Lesson 6 - Example 4.5 (p. 144)
#### Create partial regression plots for each of the two regressors. Use rstudent() to calculate residuals.

```{r}

# Delivery_Time, given Distance
model1 <- lm(Delivery_Time ~ Distance )
residuals.1 <- model1$residuals

# Num_Cases, given Distance
model2 <- lm(Num_Cases ~ Distance)
residuals.2 <- model2$residuals

# Delivery_Time, given x1
model3 <- lm(Delivery_Time ~ Num_Cases)
residuals.3 <- model3$residuals

# Distance, given Num_Cases
model4 <- lm(Distance ~ Num_Cases)
residuals.4 <- model4$residuals

plot(residuals.2, residuals.1, main='Delivery_Time|Distance ~ Num_Cases|Distance', xlab = "Number of Cases", ylab = "Residuals")
plot(residuals.4, residuals.3, main='Delivery_Time|Num_Cases ~ Distance|Num_Cases', xlab = "Distance", ylab = "Residuals")

```

## Lesson 6 - Example 4.6 (p.151-152)
#### Calculate PRESS Statistic and R-Squared_PRESS (by hand)

```{r}

# PRESS statistic using eqn 4.17
PRESS_statistic <- sum((Res/(1-diag(H)))^2)

# sum of squares total
SS_T <- t(Delivery_Time)%*%Delivery_Time-(sum(Delivery_Time))^2/length(Delivery_Time)

# R-Squared PRESS using eqn 4.18
R_Squared_PRESS <- 1 - PRESS_statistic/SS_T

```

The PRESS Statistic is $`r PRESS_statistic`$ and the $R^2$ for Prediction Based on PRESS is $R^2_{prediction}$ = $`r R_Squared_PRESS`$