---
title: "L12Ex_Voltage_Drop_Rick_Davila"
author: "Rick Davila"
date: "5/24/2020"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

#### Perform data housekeeping - upload, name columns, display to make sure it reads properly, etc. 

```{r setup, include=TRUE, results="asis"}
knitr::opts_chunk$set(echo = TRUE)

library(e1071)
library("xlsx")
library(xtable)
library(MASS) # Needed for ginv() function

rm(list = ls())

# load data
Ex72 <- read.xlsx(
  "data-ex-7-2.xlsx",
  sheetIndex = 1, sheetName=NULL, rowIndex=NULL, 
  startRow=NULL, endRow=NULL, colIndex= c(1,2,3),
  as.data.frame=TRUE, header=TRUE, colClasses=NA, 
  keepFormulas=FALSE, encoding="unknown")

# Give labels to data columns
names(Ex72) <- c("obs", 
                 "time",
                 "voltage_drop")
attach(Ex72)

# Output data to make sure it reads properly
out <- as.data.frame(c(Ex72))
colnames(out) <- c("obs", 
                 "time",
                 "voltage_drop")
tab <- (xtable(out, digits=c(0,0,1,2)))
print(tab, type="html")

# Output data structure and dimensions
str(Ex72)
dim(Ex72)

```

## Example 7.2 (p.231-234)
#### Create scatterplot

```{r}

plot(time, voltage_drop, main = "Scatterplot of time vs voltage drop",
     xlab = "time (sec)",
     ylab = "voltage drop")

```

## Fit a cubic spline using two knots, one at  t1=6.5 and one at t2=13
#### Create indicator functions for the two knots

```{r}
t1 <- 6.5
t2 <- 13

x_t1 <- ifelse((time-t1)>0,(time-t1),0)
x_t2 <- ifelse((time-t2)>0,(time-t2),0)
  
```

#### Add columns to dataframe for the higher-order terms and spline terms
 
```{r results="asis"}

Ex72$x_b01 <- time
Ex72$x_b02 <- time^2
Ex72$x_b03 <- time^3
Ex72$x_b1 <- x_t1^3
Ex72$x_b2 <- x_t2^3

# Output data to make sure it reads properly
out <- as.data.frame(c(Ex72))
#colnames(out) <- c("obs", 
#                 "time",
#                 "voltage_drop")
tab <- (xtable(out, digits=c(0,0,1,2,3,3,3,3,3)))
print(tab, type="html")

```

#### Fit model; compare to values in Table 7.4, p. 233

```{r results="asis"}

model.72 <- lm(voltage_drop ~ 
                 Ex72$x_b01 +
                 Ex72$x_b02 +
                 Ex72$x_b03 +
                 Ex72$x_b1 +
                 Ex72$x_b2)

summary(model.72)
xtable(summary(model.72))

```

#### Reproduce ANOVA table on p. 233

```{r results="asis"}

summary(model.72)
xtable(anova(model.72))

```

#### Test significance of spline terms using Partial F-test. Use alpha = 0.01 as the significance level

```{r results="asis"}

# full model including spline terms 
model.full <- lm(voltage_drop ~ 
                 Ex72$x_b01 +
                 Ex72$x_b02 +
                 Ex72$x_b03 +
                 Ex72$x_b1 +
                 Ex72$x_b2)

# reduced model excluding spline terms
model.reduced <- lm(voltage_drop ~ 
                 Ex72$x_b01 +
                 Ex72$x_b02 +
                 Ex72$x_b03)

# anova -- comparision of reduced to full model
anova(model.reduced, model.full)

# F crit
alpha <- 0.01
df_SS_R <- anova(model.reduced, model.full)$'Df'[2]
df_SS_Res <- anova(model.reduced, model.full)$'Res.Df'[2]

F_crit <- qf(1-alpha,df_SS_R,df_SS_Res)

```

We're investigating the contribution of the spline terms to the model. That is, we wish to test 
$$H_0: \beta_1=\beta_2=0 $$
we have 
$$ F_0 = \frac{SS_R(\beta_1,\beta_2|\beta_{00},\beta_{01},\beta_{02},\beta_{03}/2}{MS_{Res}}=\frac{`r anova(model.reduced, model.full)$'Sum of Sq'[2]`/`r anova(model.reduced, model.full)$'Df'[2]`}{`r anova(model.full)$'Mean Sq'[6]`}=`r anova(model.reduced, model.full)$'F'[2]` $$
and since $F_{0.01,2,35} = `r F_crit`$, we have $F_0 \gt F_{0.01,2,35}$ and as a result, we reject the null hypothesis and conclude that the spline terms contributes significantly to the model. 

#### Create residuals versus fits plot for each model

```{r}

plot(model.72$fitted.values, model.72$residuals, main = "residuals e_i, versus fitted values 天i for the cubic spline model", xlab = "天i", ylab = "e_i")
abline(0, 0, col = "gray")

# assumption, for the second model, like in the book, is to fit a cubic polynomial model and show a residuals versus fits plot

x_b01b <- time
x_b02b <- time^2
x_b03b <- time^3

model.72b <- lm(voltage_drop ~ 
                 x_b01b +
                 x_b02b +
                 x_b03b)

summary(model.72b)
xtable(summary(model.72b))

plot(model.72b$fitted.values, model.72b$residuals, main = "residuals e_i, versus fitted values 天i for the cubic polynomial model", xlab = "天i", ylab = "e_i")
abline(0, 0, col = "gray")

```



