---
title: "L12Ex_Hardwood_Conc_Rick_Davila"
author: "Rick Davila"
date: "5/23/2020"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

#### Perform data housekeeping - upload, name columns, display to make sure it reads properly, etc. 

```{r setup, include=TRUE, results="asis"}
knitr::opts_chunk$set(echo = TRUE)

library(e1071)
library("xlsx")
library(xtable)
library(MASS) # Needed for ginv() function

rm(list = ls())

# load data
Ex71 <- read.xlsx("data-ex-7-1.xlsx",
  sheetIndex = 1, sheetName=NULL, rowIndex=NULL, 
  startRow=NULL, endRow=NULL, colIndex= c(1,2),
  as.data.frame=TRUE, header=TRUE, colClasses=NA, 
  keepFormulas=FALSE, encoding="unknown")

# Give labels to data columns
names(Ex71) <- c("Concentration", 
                 "Strength")
attach(Ex71)

# Output data to make sure it reads properly
out <- as.data.frame(c(Ex71))
colnames(out) <- c("Concentration", 
                 "Strength")
tab <- (xtable(out, digits=c(0,1,1)))
print(tab, type="html")

# Output data structure and dimensions
str(Ex71)
dim(Ex71)

```

## Example 7.1 (p. 227-229)
#### Create scatterplot

```{r}

plot(Concentration, Strength, main = "Scatterplot of Hardwood Concentration vs Tensile Strength",
     xlab = "hardwood concentration (%)",
     ylab = "tensile strength")

```

#### Fit a quadratic model on centered data; compare to p. 227

```{r results="asis"}

n <- length(Strength)
x_mean <- sum(Concentration)/n
x_1 <- Concentration - x_mean
x_2 <- x_1^2

model.71 <- lm(Strength ~ x_1 + x_2)

summary(model.71)
xtable(summary(model.71))

```

#### Create ANOVA table (by hand and with R); compare to Table 7.2 on p. 228

```{r results="asis"}

# create ANOVA table by hand
n <- length(Strength)
k <- 2                    # number of regressors
p <- k + 1                # number of parameters

X <- cbind(matrix(1,length(Strength),1),as.matrix(x_1),as.matrix(x_2))
y <- as.matrix(Strength)

# X'X matrix
xTx <- t(X) %*% X

# Beta Coefficients
beta_hat <- ginv(xTx,tol =  .Machine$double.eps) %*% t(X) %*% y

# eqn 3.24
SS_Reg <- t(beta_hat) %*% t(X) %*% y - (sum(y))^2/n 

MS_Reg <- SS_Reg/k

# eqn 3.16
SS_Res <- t(y) %*% y - t(beta_hat) %*% t(X) %*% y

# eqn 3.17
MS_Res <- SS_Res/(n-p)

# eqn 3.24
SS_T <- t(y) %*%  y - (sum(y))^2/n 

F_0 <- MS_Reg/MS_Res

# t-critical and p-value calculations
# for alpha = 0.05
alpha = 0.05

p_val_f = 1 - pf(F_0, k, n-p)

# create ANOVA table with R
xtable(anova(model.71))

```

Anova table by hand is listed above.  The anova table by hand (i.e., values) are listed below:

SS Regression: $`r SS_Reg`$

SS Residual: $`r SS_Res`$

SS Total: $`r SS_Res + SS_Reg`$

Regression Degrees of Freedom: $`r k`$

Residual Degrees of Freedom: $`r n-p`$   

Total Degrees of Freedom: $`r k+n-p`$

Mean Square Regression: $`r MS_Reg`$

Mean Square Residual: $`r MS_Res`$

$F_0 = `r F_0`$

P Value: $`r p_val_f`$


#### Test significance of quadratic term using Partial F-test. Use alpha = 0.01 as the significance level. Compare to p. 229

```{r results="asis"}

# full model including the quadratic term 
model.full <- lm(Strength ~ x_1 + x_2)

# reduced model excluding the quadratic term
model.reduced <- lm(Strength ~ x_1)

# anova -- comparision of reduced to full model
anova(model.reduced, model.full)

# F crit
alpha <- 0.01
df_SS_R <- anova(model.reduced, model.full)$'Df'[2]
df_SS_Res <- anova(model.reduced, model.full)$'Res.Df'[2]

F_crit <- qf(1-alpha,df_SS_R,df_SS_Res)

```

We're investigating the contribution of the quadratmodel.reduced, model.fullic term to the model. That is, we wish to test $$H_0: \beta_2=0, H_1:\beta_2 \ne 0 $$  we have $$ F_0 = \frac{SS_R(\beta_2|\beta_1,\beta_0)/1}{MS_{Res}}=\frac{`r anova(model.reduced, model.full)$'Sum of Sq'[2]`/`r anova(model.reduced, model.full)$'Df'[2]`}{`r anova(model.full)$'Mean Sq'[3]`}=`r anova(model.full)$'F'[2]` $$ and since $F_{0.01,1,16} = `r F_crit`$, we have $F_0 \gt F_{0.01,1,16}$ and as a result, we conclude that the quadratic term contributes significantly to the model. 

#### Create residuals versus fits plot

```{r}

plot(model.71$fitted.values, model.71$residuals, main = "residuals (e_i) versus fitted values (ŷ_i)", xlab = "ŷi", ylab = "e_i")
abline(0, 0, col = "gray")

```

This plot does not reveal any serious model inadequacy.

#### Create normal probability plot of residuals

```{r}

# set datax = TRUE to resemble plot in book; default value is datax = FALSE
# however, y- and x-label won't change iaw datax = TRUE
qqnorm(model.71$residuals, datax = TRUE, main="normal probability plot of the residuals", ylab = "e_i", xlab = "expanded normal value")
abline(0, 0, col = "gray")
qqline(model.71$residuals, datax = TRUE)

```

The normal probability plot of the residuals indicates that the error distribution has heavier tails than the normal.
